---
title: "Figure 2"
output:
 html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
date: "2024-09-11"
---

# Setup
```{r setup, include=FALSE}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#library(ropls)

#install.packages("pls")

library(googledrive)
library(here)

#library(pls)
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(GenomicFeatures)
library(tidyverse)


#  BiocManager::install('PCAtools')
library(PCAtools)
library(patchwork)
library(gridExtra)


select = dplyr::select


types = c("Stipe","Lamella","Spore","control")
colors = c("#96c176ff","#cf9ddbff","#4472C4ff","#000000FF")
types = c("Stipe","Lamella","Spore","control","stipe","lamella","spore")
colors = c("#96c176ff","#cf9ddbff","#4472C4ff","#000000FF","#96c176ff","#cf9ddbff","#4472C4ff")
names(colors) = types
  

annotationColors = c("mRNA" = "#4472C4","TE" = "#C00000","Both" = "#9D475B","intergenic" = "#AFABAB")



methylationLevelColors = my_colors <- c(
  "Nothing"   = "#4575b4",
  "Low"  = "#91bfdb",
  "Medium"  = "#e0f3f8",
  "High"  = "#fee090",
  "All" = "#d73027"
)



```




# Annotation



### TE reference
```{r, message=FALSE}
gff3Colnames =c("seqid", "source",
                "type","start",
                "end", "score",
                "strand","phase",
                "attributes")


  TE.classes = read_csv(file = here("annotation","Repeat.classes.csv")) %>%
    rename(TEfamily = "Name") %>%
    rename(Notes = "Superfamily / Notes")

  
  TEFile <- here("annotation","Maror2.repeatmasker.rmlib2.3.gff")

  TEdf =    read_tsv(file = TEFile, col_names = gff3Colnames,comment = "#") %>% separate(attributes, into = c("ID","target"), sep = ";") %>%
    mutate(ID = as.numeric(gsub("ID=","",ID))) %>%
    separate(target, into = c("Target","motif",
                              "motifStart","motifStop"),
             sep = " ") %>%
    mutate(motif = gsub("\"","",motif))%>%
    mutate(motif = gsub("Motif:","",motif))  %>%
    select(seqid,start,end,score,
           strand, ID,motif,
           motifStart,motifStop) %>%
    rename(TEID = "ID") %>%
    mutate(TE_ID2 = 1:n()) %>%
    mutate(motifStart = as.numeric(gsub(" ","",motifStart))) %>%
    mutate(motifStop = as.numeric(gsub(" ","",motifStop))) %>%
    dplyr::rename(TEclass = "motif") %>%
    mutate(TEclass = gsub(" ","",TEclass)) %>%
    mutate(TEclass = gsub("Maror","",TEclass)) 

 

  TEs = makeGRangesFromDataFrame(TEdf,
                                 seqnames.field = "seqid", 
                                 start.field = "start", 
                                 end.field = "end",
                                 strand.field = "strand",
                                 keep.extra.columns = TRUE)
  

TEdf%>%
  separate(TEclass, into = c("TEfamily","nr"), sep = "-",remove = FALSE) %>%
  inner_join(TE.classes)->
  TEdf

TEdf %>%
  group_by(TEfamily) %>% summarise(n = n(), min = min(motifStart), max = max(motifStop)) %>%
  mutate(TElength = max - min) %>%
  filter(TElength > 1000) %>%
  inner_join(TEdf) %>%
  mutate(fraction =(motifStop-motifStart) / TElength)->
  TEdf.long

TEdf.long  %>% separate(TEclass, into = c("TEfamily","nr"), sep = "-",remove = FALSE) ->
  TEdf.long

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(grepl("DNA",Class)) %>%
  ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(Class~TEfamily, scales = "free_x", space = "free") -> 
  TE.distribution.ClassII.plot

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(grepl("Retro",Class)) %>%
    ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(Class~TEfamily, scales = "free_x", space = "free") -> 
 TE.distribution.ClassI.plot

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(!grepl("DNA",Class)) %>%
  filter(!grepl("Retro",Class)) %>%
  ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(.~TEfamily, scales = "free_x", space = "free") -> 
 TE.distribution.Other.plot


TE.distribution.ClassI.plot+
  TE.distribution.ClassII.plot+
  TE.distribution.Other.plot +
  plot_annotation(tag_levels = 'A')+plot_layout(ncol = 1) ->
  TE.distribution.plots

#ggsave(here("results","annotation","TE.distribution.plots.pdf"),
#       TE.distribution.plots,
#       width = 10, height = 15, units = "in", dpi = 300)

  
TEdf.long %>% arrange(TEclass, -fraction) %>% 
  filter(fraction > 0.25)

```


## Gene reference 
```{r, message=FALSE}

gtfFile = here("annotation","GCF_018924745.1_UU_Maror_2_genomic.gff")	
txdb <- makeTxDbFromGFF(file=gtfFile, dataSource="ensemblgenomes",format = "gff3")


gff3Colnames =c("seqid", "source",
                "type","start",
                "end", "score",
                "strand","phase",
                "attributes")

  annotation.df =    read_tsv(file = gtfFile, col_names = gff3Colnames,comment = "#") 


 annotation.df %>% filter(type =="mRNA") %>%
   group_by(seqid,start,end, strand) %>%
   slice_head(n = 1) %>%
   ungroup() %>%
   separate_longer_delim(attributes, delim = ";") ->
   test
 
  test %>% select(seqid,start,end,strand,attributes) %>%
   separate(attributes, into = c("key","value"), sep = "=") %>%
   filter(key %in% c("Name","locus_tag","product")) %>%
   pivot_wider(values_from = "value",names_from = "key") -> 
   annotation.mRNA.df

 annotation.mRNA.df %>% group_by(product) %>% summarise(n = n())
 
annotation.mRNA.df %>% select(seqid,start,end, strand,Name,locus_tag,product) %>%
  mutate(tx_id = 1:n()) -> 
  annotation.mRNA.df

mRNAs = makeGRangesFromDataFrame(annotation.mRNA.df,
                                 seqnames.field = "seqid", 
                                 start.field = "start", 
                                 end.field = "end",
                                 strand.field = "strand",
                                 keep.extra.columns = TRUE)
  

```

Run only once for getting relationship between protein ID (XP) and gene ID (XM)
```{r, inlcude false  }

gene2ProteinFile = here("annotation","gene2protein_relationship.csv")

relationship_df = read_csv(gene2ProteinFile)

relationship_df %>% mutate(Name = XM_ID) -> relationship_df
annotation.mRNA.df = inner_join(annotation.mRNA.df,relationship_df , by = "Name")
```

## Distance between RNA and TEs
```{r, message=FALSE}
# Find the nearest TE for each mRNA
# query = mRNAs, subject = TEs
nearest_hits <- distanceToNearest(mRNAs, TEs, ignore.strand=TRUE)
# Extract results into a readable format
# This creates a data frame with mRNA index, TE index, and the distance
results <- as.data.frame(nearest_hits)

# Optional: Merge metadata to see which mRNA is near which TE
results$locus_tag <- mRNAs$locus_tag[results$queryHits]
results$TE_ID2   <- TEs$TE_ID2[results$subjectHits] # Replace 'name' with your TE column name

results %>%
  select(locus_tag, TE_ID2, distance) ->
  GeneBody_TE_distance

```






# Samples 
```{r, message=FALSE}

select = dplyr::select

samples.ring.methyl.df = read_csv(file = here("data","meDNA","meDNA.metaData.csv"))


samples.ring.methyl.df %>% arrange(Tissue,Location,Replicate) -> 
  samples.ring.methyl.df


samples.ring.methyl.df %>% group_by(Tissue,Location) %>% summarise(n = n()) %>%
  pivot_wider(names_from = "Location", values_from = "n")->
  samples.ring.methyl.df.summary



# make the directory if it does not exist
dir.create(here("results","meDNA"), showWarnings = FALSE, recursive = TRUE)

# Open the PDF device
pdf(here("results","meDNA","meDNA.samples.summary"), width = 8, height = 11)
# Draw the table
grid.table(samples.ring.methyl.df.summary)
# Close the device
dev.off()

```




# Read in data

Loading it from the google drive. This might change when we publish

```{r, message=FALSE}

# This will prompt you to log in to your Google account in your browser
drive_auth()

# Read the file using its ID or name
file_id <- as_id("1C9qPmUr6oE80PXn8adjG1q4-pRvqQkRk")

# 2. Download to a temporary location with the correct extension
# This ensures read_csv knows it's a gzipped file
temp_path <- tempfile(fileext = ".csv.gz")
drive_download(file_id, path = temp_path, overwrite = TRUE)

# 3. Read it - read_csv automatically handles the decompression
methylationData.3.QC <- read_csv(temp_path)


```


# Figure2

Change to Intergenic (Capital) (DONE)
Change to fraction instead of percent (DONE)
Change to Gene body instead of GeneBody (Not done make in ai)
Make a statistical test for the differences in methyalation level.



```{r, message=FALSE}


methylationData.3.QC %>% 
  group_by(location,Tissue,Location,Replicate2) %>% 
  summarise(n = n(), depth = mean(depth), methylation = mean(percent)) ->
  methylationData.3.QC.annotation.distribution


 

methylationData.3.QC.annotation.distribution %>%
  filter(location != "Both") %>%
  select(Tissue,Location,Replicate2,methylation,location) %>%
  mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
  mutate(fraction = methylation /100)%>%
  ggplot(aes(x = Tissue, y = fraction, fill = Tissue)) + # Change x to Tissue
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.2) + 
#  geom_text(aes(label =IDsample ))+ # Add alpha for better visualization if many points
  theme_light() +
  scale_fill_manual(values = colors) +
  facet_wrap(.~location, scales = "free_y") -> fig2C

ggsave(here("figures","Figure2","fig2C.methylationData.samples.annotation.distribution.1.pdf"),
            plot = fig2C,width = 8, height = 5, units = "in", dpi = 300)



methylationData.3.QC.annotation.distribution %>%
  filter(location != "Both") %>%
  select(Tissue,Location,Replicate2,methylation,location) %>%
  mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
  mutate(fraction = methylation /100)%>%
  ggplot(aes(x = location, y = fraction, fill = location)) + # Change x to Tissue
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.2) + 
#  geom_text(aes(label =IDsample ))+ # Add alpha for better visualization if many points
  theme_light() +
  ylim(c(0,1))+
  scale_fill_manual(values = colors)  -> fig2A

ggsave(here("figures","Figure2","fig2A.methylationData.samples.annotation.distribution.1.pdf"),
       plot = fig2A,
        width = 4, height = 5, units = "in", dpi = 300)


fig2A

```


```{r, message=FALSE}

methylationData.3.QC %>% 
  group_by(Tissue,Location,Replicate) %>% 
  summarise(n = n(), depth = mean(depth), methylation = mean(percent)) ->
  methylationData.3.QC.distribution


methylationData.3.QC.distribution %>% 
  mutate(IDsample = paste(Tissue,Location,Replicate,sep = "_")) %>%
  ggplot(aes(x = IDsample, y = depth, fill = Tissue),color = "grey") +
  geom_hline(yintercept = 10, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5)) 
  
  

methylationData.3.QC.annotation.distribution %>% 
  dplyr::rename(annotation = "location") %>%
  mutate(IDsample = paste(Tissue,Location,Replicate2,sep = "_")) %>%
  ggplot(aes(x = IDsample, y = depth, fill = annotation),color = "grey") +
  geom_bar(stat = "identity",position = "fill") +
  scale_fill_manual(values = annotationColors) + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5)) ->
  distribution.depth

methylationData.3.QC.annotation.distribution %>% 
  dplyr::rename(annotation = "location") %>%
  mutate(IDsample = paste(Tissue,Location,Replicate2,sep = "_")) %>%
  ggplot(aes(x = IDsample, y = methylation, fill = annotation),color = "grey") +
  geom_bar(stat = "identity",position = "dodge") +
  scale_fill_manual(values = annotationColors) + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5)) 




methylationData.3.QC %>% group_by(Tissue,Location,Replicate) %>%
  summarise(sum = sum(n())) %>% 
  inner_join(methylationData.3.QC.annotation.distribution) %>%
  mutate(fraction = n / sum) %>%
  mutate(location = factor(location, levels = c("mRNA","Both","TE","intergenic"))) %>%
  dplyr::rename(annotation = "location") %>%
  ggplot(aes(x = Tissue, y = n, fill = annotation)) +geom_bar(stat = "identity", color = "black") +
  facet_wrap("Location")+ 
  scale_fill_manual(values = annotationColors) 
#ggsave("../../results/meDNA/locationDistribution.pdf")

 

  methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    pivot_wider(values_from = "methylation", names_from = "location" )  %>%
    ggplot(aes(x = GeneBody, y = TE, color = Tissue)) + geom_point()+
    theme_light()+
    geom_text_repel(aes(label = IDsample), color = "black")  +
    scale_color_manual(values = colors) -> sampleDistribution
#ggsave("../../results/meDNA/sampleDistribution.pdf")
  
methylationData.3.QC.annotation.distribution %>% 
  select(Tissue,Location,Replicate2,methylation,location) %>%
  mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
  filter(IDsample %in%
           c("northeast_1_stipe","north_3_stipe","north_3_spore")) %>%
    mutate(fraction = methylation /100) %>%
    select(-methylation) %>%
    pivot_wider(values_from = "fraction", names_from = "location" ) ->outliers  
  

  methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    mutate(fraction = methylation /100) %>%
    select(-methylation) %>%
    pivot_wider(values_from = "fraction", names_from = "location" )  %>%
    ggplot(aes(x = GeneBody, y = TE, color = Tissue)) + geom_point()+
    theme_light()+
    geom_text_repel(data = outliers,aes(label = IDsample), color = "black")  +
    scale_color_manual(values = colors) -> 
    fig2B
  
  fig2B
ggsave(here("figures","Figure2","Fig2B.TE.GeneBody.scatterPlot.pdf"),plot = fig2B,
       width = 6, height = 5, units = "in", dpi = 300)


methylationData.3.QC %>%
  mutate(methylated = depth*percent /100) %>%
  mutate(notMethylated = depth - methylated) %>%
  group_by(Sample) %>%
  summarise(methylated = sum(methylated), notMethylated = sum(notMethylated)) %>%
  mutate(fraction = methylated /(notMethylated+methylated))->
  sum

sum %>%
  inner_join(samples.ring.methyl.df) %>%
  ggplot(aes(x = Tissue, y = fraction,fill = Tissue,)) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(width = 0.1,mapping = aes(shape =Location)) + 
  theme_light()+
  scale_fill_manual(values = colors) ->
  averageMethylation


methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    mutate(fraction = methylation /100) %>%
  group_by(location) %>%
  summarise(
    p_value = tryCatch({
      kruskal.test(fraction ~ Tissue, data = cur_data())$p.value
    }, error = function(e) NA),
  )%>% 
  mutate(comparison = "Location") %>%
  mutate(adj_pvalue = p.adjust(p_value, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.05), "Yes")) ->
  kruskal_results.Methylation


library(rstatix)
library(purrr)

results <- methylationData.3.QC.annotation.distribution %>% 
  select(Tissue, Location, Replicate2, methylation, location) %>%
  mutate(
    IDsample = paste(Location, Replicate2, Tissue, sep = "_"),
    fraction = methylation / 100
  ) %>%
  group_by(location) %>%
  group_modify(~{
    
    df <- .x  # current location
    
    ## --- Kruskal–Wallis test ---
    kw <- tryCatch(
      kruskal.test(fraction ~ Tissue, data = df),
      error = function(e) NULL
    )
    
    kw_p <- if (!is.null(kw)) kw$p.value else NA
    
    ## --- Dunn test (pairwise comparisons) ---
    dunn <- df %>%
      dunn_test(fraction ~ Tissue, p.adjust.method = "BH") %>%
      mutate(
        kw_p_value = kw_p
      )
    
    dunn
  }) %>%
  ungroup() %>%
  
  ## Adjust the Kw p-values across all locations
  group_by(location) %>%
  mutate(
    kw_p_adj = p.adjust(unique(kw_p_value), method = "BH")
  ) %>%
  ungroup() %>%
  
  ## Add significance flags
  mutate(
    kw_significant = kw_p_adj < 0.05,
    pairwise_significant = p.adj < 0.05
  )

write_csv(results, here("figures","Figure2","fig2C.methylationData.samples.annotation.distribution.kruskal.dunn.test.results.csv"))

```




```{r, message=FALSE}


methylationData.3.QC %>% 
  group_by(Tissue, location) %>%
  summarise(mean = mean(percent))
  
methylationData.3.QC %>% 
  ggplot(aes(x = percent,color = Tissue )) +
  geom_density()+
  geom_vline(xintercept = 90, linetype = "dashed", color = "red")+
  geom_vline(xintercept = 10, linetype = "dashed", color = "blue")+
  scale_color_manual(values = colors)  +
  theme_minimal() +
  labs(
    fill = "Methylation Level", 
    x = "Precent methylated per site", 
    y = "Density"
  )->methylation.density.plot
  



methylationData.3.QC %>% 
  mutate(methylationLevel = "High") %>%
  mutate(methylationLevel = replace(methylationLevel,
                                    percent <90,
                                    "Average" )) %>%
  mutate(methylationLevel = replace(methylationLevel,
                                    percent <10,
                                    "Low" )) ->
  methylationData.3.QC






methylationData.3.QC %>% 
  group_by(Location,location, Tissue,methylationLevel) %>%
  summarise(n = n()) -> methylationData.3.QC.summary

methylationData.3.QC.summary %>%
  group_by(Location,location, Tissue) %>%
  summarise(Total = sum(n)) %>% 
  inner_join(methylationData.3.QC.summary) %>%
  mutate(fraction = n /Total) -> methylationData.3.QC.summary

methylationData.3.QC <- methylationData.3.QC %>%
  mutate(methylationLevel = factor(methylationLevel, levels = c("High", "Average", "Low")))


ggplot(methylationData.3.QC, aes(x = Tissue, fill = methylationLevel)) +
  geom_bar(position = "fill") +
  facet_wrap(facets = vars(location)) + # Ensure 'location' is used properly
  scale_fill_manual(values = c("Low" = "blue", "Average" = "grey", "High" = "red")) +
  theme_minimal() +
  labs(
    fill = "Methylation Level", 
    x = "Tissue", 
    y = "Count"
  )-> methylationDistribution
ggsave(here("figures","Figure2","MethylationDistribution.pdf"),plot = methylationDistribution)

methylationDistribution

layout= c("AB
          CC
          CC")
#methylation.density.plot+ methylationDistribution+
#    sampleDistribution + plot_layout(design = layout, guides = "collect") +
#  plot_annotation(tag_levels = 'A')-> figure2
#ggsave("~/MarasmiusMethylationRNA/suppFigures/Figure2.methylation.pdf",width = 10, height = 8, units = "in", dpi = 300,plot = figure2)

#ggsave("~/MarasmiusMethylationRNA/figures/Figure2.methylation.annotation.png",
#       plot = figure2, width = 180,
#       height = 180, units = "mm", 
#       dpi = 300,plot = figure2)


```





