---
title: "Methylation pattern"
output: html_document
date: "2024-09-11"
---

# Setup
```{r setup, include=FALSE}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#library(ropls)

#install.packages("pls")

library(googledrive)
library(here)

#library(pls)
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(GenomicFeatures)
library(tidyverse)


#  BiocManager::install('PCAtools')
library(PCAtools)
library(patchwork)
library(gridExtra)


# For linear model
library(lme4)
library(lmerTest)   # adds p-values


# For handling fasta files
library(Biostrings)


select = dplyr::select


types = c("Stipe","Lamella","Spore","control")
colors = c("#96c176ff","#cf9ddbff","#4472C4ff","#000000FF")
types = c("Stipe","Lamella","Spore","control","stipe","lamella","spore")
colors = c("#96c176ff","#cf9ddbff","#4472C4ff","#000000FF","#96c176ff","#cf9ddbff","#4472C4ff")
names(colors) = types
  

annotationColors = c("mRNA" = "#4472C4","TE" = "#C00000","Both" = "#9D475B","intergenic" = "#AFABAB")



methylationLevelColors = my_colors <- c(
  "Nothing"   = "#4575b4",
  "Low"  = "#91bfdb",
  "Medium"  = "#e0f3f8",
  "High"  = "#fee090",
  "All" = "#d73027"
)



```




# Annotation


### TE fastaReference
```{r, message=FALSE}

# Load the TE fasta file
TEfastaFile = here("reference","Maror2.repeat.library.2.3.fasta")
TEfasta = readDNAStringSet(TEfastaFile)

# remove all the - in the sequences
TEfasta = DNAStringSet(gsub("-", "", as.character(TEfasta)))

# Get all the lengths in a data frame
TEfasta.df = data.frame(
  TE_ID = names(TEfasta),
  length = width(TEfasta)) %>%
  mutate(TE_ID = gsub("Maror","",TE_ID)) %>%
  separate(TE_ID, into = c("TEclass","other"), sep = "/") %>%
  separate(TEclass, into = c("TEclass","class"), sep = "#") %>%
   separate(TEclass, into = c("TEfamily","nr"), sep = "-", remove = FALSE) %>% 
  mutate(nr = as.numeric(nr)) %>%
  select(-other)
  
  

```







### TE reference
```{r, message=FALSE}
gff3Colnames =c("seqid", "source",
                "type","start",
                "end", "score",
                "strand","phase",
                "attributes")


  TE.classes = read_csv(file = here("annotation","Repeat.classes.csv")) %>%
    rename(TEfamily = "Name") %>%
    rename(Notes = "Superfamily / Notes")

  
  TEFile <- here("annotation","Maror2.repeatmasker.rmlib2.3.gff")

  TEdf =    read_tsv(file = TEFile, col_names = gff3Colnames,comment = "#") %>% separate(attributes, into = c("ID","target"), sep = ";") %>%
    mutate(ID = as.numeric(gsub("ID=","",ID))) %>%
    separate(target, into = c("Target","motif",
                              "motifStart","motifStop"),
             sep = " ") %>%
    mutate(motif = gsub("\"","",motif))%>%
    mutate(motif = gsub("Motif:","",motif))  %>%
    select(seqid,start,end,score,
           strand, ID,motif,
           motifStart,motifStop) %>%
    rename(TEID = "ID") %>%
    mutate(TE_ID2 = 1:n()) %>%
    mutate(motifStart = as.numeric(gsub(" ","",motifStart))) %>%
    mutate(motifStop = as.numeric(gsub(" ","",motifStop))) %>%
    dplyr::rename(TEclass = "motif") %>%
    mutate(TEclass = gsub(" ","",TEclass)) %>%
    mutate(TEclass = gsub("Maror","",TEclass)) 

 

  TEs = makeGRangesFromDataFrame(TEdf,
                                 seqnames.field = "seqid", 
                                 start.field = "start", 
                                 end.field = "end",
                                 strand.field = "strand",
                                 keep.extra.columns = TRUE)
  

TEdf%>%
  separate(TEclass, into = c("TEfamily","nr"), sep = "-",remove = FALSE) %>%
  inner_join(TE.classes)->
  TEdf

TEdf %>%
  group_by(TEfamily) %>% summarise(n = n(), min = min(motifStart), max = max(motifStop)) %>%
  mutate(TElength = max - min) %>%
  filter(TElength > 1000) %>%
  inner_join(TEdf) %>%
  mutate(fraction =(motifStop-motifStart) / TElength)->
  TEdf.long

TEdf.long  %>% separate(TEclass, into = c("TEfamily","nr"), sep = "-",remove = FALSE) ->
  TEdf.long

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(grepl("DNA",Class)) %>%
  ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(Class~TEfamily, scales = "free_x", space = "free") -> 
  TE.distribution.ClassII.plot

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(grepl("Retro",Class)) %>%
    ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(Class~TEfamily, scales = "free_x", space = "free") -> 
 TE.distribution.ClassI.plot

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(!grepl("DNA",Class)) %>%
  filter(!grepl("Retro",Class)) %>%
  ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(.~TEfamily, scales = "free_x", space = "free") -> 
 TE.distribution.Other.plot


TE.distribution.ClassI.plot+
  TE.distribution.ClassII.plot+
  TE.distribution.Other.plot +
  plot_annotation(tag_levels = 'A')+plot_layout(ncol = 1) ->
  TE.distribution.plots

#ggsave(here("results","annotation","TE.distribution.plots.pdf"),
#       TE.distribution.plots,
#       width = 10, height = 15, units = "in", dpi = 300)

  
TEdf.long %>% arrange(TEclass, -fraction) %>% 
  filter(fraction > 0.25)

```


## Gene reference 
```{r, message=FALSE}

gtfFile = here("annotation","GCF_018924745.1_UU_Maror_2_genomic.gff")	
txdb <- makeTxDbFromGFF(file=gtfFile, dataSource="ensemblgenomes",format = "gff3")


gff3Colnames =c("seqid", "source",
                "type","start",
                "end", "score",
                "strand","phase",
                "attributes")

  annotation.df =    read_tsv(file = gtfFile, col_names = gff3Colnames,comment = "#") 


 annotation.df %>% filter(type =="mRNA") %>%
   group_by(seqid,start,end, strand) %>%
   slice_head(n = 1) %>%
   ungroup() %>%
   separate_longer_delim(attributes, delim = ";") ->
   test
 
  test %>% select(seqid,start,end,strand,attributes) %>%
   separate(attributes, into = c("key","value"), sep = "=") %>%
   filter(key %in% c("Name","locus_tag","product")) %>%
   pivot_wider(values_from = "value",names_from = "key") -> 
   annotation.mRNA.df

 annotation.mRNA.df %>% group_by(product) %>% summarise(n = n())
 
annotation.mRNA.df %>% select(seqid,start,end, strand,Name,locus_tag,product) %>%
  mutate(tx_id = 1:n()) -> 
  annotation.mRNA.df

mRNAs = makeGRangesFromDataFrame(annotation.mRNA.df,
                                 seqnames.field = "seqid", 
                                 start.field = "start", 
                                 end.field = "end",
                                 strand.field = "strand",
                                 keep.extra.columns = TRUE)
  

```

Run only once for getting relationship between protein ID (XP) and gene ID (XM)
```{r, inlcude false  }

gene2ProteinFile = here("annotation","gene2protein_relationship.csv")

relationship_df = read_csv(gene2ProteinFile)

relationship_df %>% mutate(Name = XM_ID) -> relationship_df
annotation.mRNA.df = inner_join(annotation.mRNA.df,relationship_df , by = "Name")
```

## Distance between RNA and TEs
```{r, message=FALSE}
# Find the nearest TE for each mRNA
# query = mRNAs, subject = TEs
nearest_hits <- distanceToNearest(mRNAs, TEs, ignore.strand=TRUE)
# Extract results into a readable format
# This creates a data frame with mRNA index, TE index, and the distance
results <- as.data.frame(nearest_hits)

# Optional: Merge metadata to see which mRNA is near which TE
results$locus_tag <- mRNAs$locus_tag[results$queryHits]
results$TE_ID2   <- TEs$TE_ID2[results$subjectHits] # Replace 'name' with your TE column name

results %>%
  select(locus_tag, TE_ID2, distance) ->
  GeneBody_TE_distance

```






# Samples 
```{r, message=FALSE}

select = dplyr::select

samples.ring.methyl.df = read_csv(file = here("data","meDNA","meDNA.metaData.csv"))


samples.ring.methyl.df %>% arrange(Tissue,Location,Replicate) -> 
  samples.ring.methyl.df


samples.ring.methyl.df %>% group_by(Tissue,Location) %>% summarise(n = n()) %>%
  pivot_wider(names_from = "Location", values_from = "n")->
  samples.ring.methyl.df.summary



# make the directory if it does not exist
dir.create(here("results","meDNA"), showWarnings = FALSE, recursive = TRUE)

# Open the PDF device
pdf(here("results","meDNA","meDNA.samples.summary"), width = 8, height = 11)
# Draw the table
grid.table(samples.ring.methyl.df.summary)
# Close the device
dev.off()

```




# Read in data

Loading it from the google drive. This might change when we publish

```{r, message=FALSE}

# This will prompt you to log in to your Google account in your browser
drive_auth()

# Read the file using its ID or name
file_id <- as_id("1C9qPmUr6oE80PXn8adjG1q4-pRvqQkRk")

# 2. Download to a temporary location with the correct extension
# This ensures read_csv knows it's a gzipped file
temp_path <- tempfile(fileext = ".csv.gz")
drive_download(file_id, path = temp_path, overwrite = TRUE)

# 3. Read it - read_csv automatically handles the decompression
methylationData.3.QC <- read_csv(temp_path)


```



#Figure 3

Check that the mRNA and TE analysis are similar in terms of significant sites

###  TE analysis

```{r }
TEdf %>% 
  mutate(motifStart = as.numeric(motifStart)) %>%
  mutate(motifStop = as.numeric(motifStop)) %>%
  separate(TEclass, into=c("TEclass","nr"), sep = "-") ->
  TEdf
TEdf %>%
  group_by(TEclass) %>%
  summarise(motifStart = min(motifStart), motifStop = max(motifStop)) %>%
  arrange(-motifStop)  %>%
  filter(motifStop>500) %>%
  select(TEclass) %>%
  inner_join(TEdf)->
  TEdf.long

TEdf.long %>% select(TEclass,seqid,TE_ID2) -> 
  TE.info

TE.info 
  


methylationData.3.QC %>% 
  select(-TEclass) %>%
  filter(location == "TE") %>%
  inner_join(TE.info) ->
  methylationData.3.QC.TE
```


####PCA analysis

```{r }
methylationData.3.QC.TE %>%
  group_by(TE_ID2,Location,Replicate2,Tissue) %>%
  summarise( percent = mean(percent)) ->
  methylationData.3.QC.TE.summary


methylationData.3.QC.TE.summary %>%
  group_by(TE_ID2,Tissue) %>%
  summarise(nrOfExamples = n()) -> methylationData.3.QC.TE.summary.2

methylationData.3.QC.TE.summary.2 %>%
  group_by(Tissue) %>%
  summarise(n = max(nrOfExamples)) %>%
  inner_join(methylationData.3.QC.TE.summary.2) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n == 3)  %>%
  select(TE_ID2)->
  TE.covered


TE.covered
  

methylationData.3.QC.TE.summary %>%
  inner_join(TE.covered) %>%
  pivot_wider(values_from = "percent",names_from = c("Location","Replicate2","Tissue")) %>%
  column_to_rownames(var = "TE_ID2")%>%
  drop_na()->                     # Removes any row with at least one NA->
  methylationData.3.QC.TE.wide


## PCA analysis
PCAresult = PCAtools::pca(methylationData.3.QC.TE.wide,center = TRUE)
rot = PCAresult$rotated[,1:6]

## Cluster analysis
dist_matrix <- dist(rot, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
library(NbClust)
nb <- NbClust(rot, distance="euclidean", min.nc=2, max.nc=10, method="ward.D2")
nb$Best.nc


wss <- numeric(10)
for (k in 2:10) {
  clusters <- cutree(hc, k)
  wss[k] <- sum(sapply(unique(clusters), function(c) {
    pts <- rot[clusters == c, ]
    sum(scale(pts, scale=FALSE)^2)
  }))
}

plot(1:10, wss, type="b", xlab="k", ylab="Within-cluster SS")


## Variance
PCA.stats = data.frame(variance = PCAresult$variance, 
                       PC = factor(names(PCAresult$variance), levels = names(PCAresult$variance)))
ggplot(PCA.stats, aes(x = PC, y = variance) )+
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(
    title = "PC variance",
    x = "PC",
    y = "Variance"
  )+theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  ) -> variation

ggsave(here("figures","Figure3","TE.PCA.variance.plot.pdf"),
       plot = variation, width = 10, height = 8, units = "in")


methylationData.3.QC.TE %>%
  group_by(Location,Replicate2,Tissue) %>%
  summarise(n = n(),mean = mean(mean)) %>%
  mutate(sampleID = paste(Location,Replicate2,Tissue, sep = "_")) ->
  sampleInfo.TE.summary

sampleInfo.TE.summary

PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary) %>%
  ggplot(aes(x = PC1, y = PC2, color =Tissue, shape = Location )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  scale_color_manual(values = colors) +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  )-> PCA.plot


  
PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary)  %>%
  ggplot(aes(x = PC1, y = PC2, color =Tissue, shape = Location,label =sampleID )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  scale_color_manual(values = colors) +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  ) + geom_text_repel() -> PCA.plot.labels



PCA.plot.labels


breaks1 <- seq(from = -100, to = 300, length.out = 4)
library(RColorBrewer)
colors1 = colorRampPalette(rev(brewer.pal(n = 7, name =
  "RdYlBu")))(4)
myColor2 <-circlize::colorRamp2(breaks = breaks1,colors =colors1)


library(ComplexHeatmap)

pdf("~/MarasmiusMethylationRNA/figures/Figure3/TE.methylation.PCA.heatmap.pdf",
    width = 8, height = 6)
ComplexHeatmap::Heatmap(
  PCAresult$rotated[1:3] ,
  col = myColor2,
  column_split = 6,
  split = clusters,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(
      x, y, width = width, height = height,
      gp = gpar(col = "darkgrey", fill = NA, lwd = 1) # Add gray borders
    )
  }
) 
dev.off()


png("~/MarasmiusMethylationRNA/figures/Figure3/TE.methylation.PCA.heatmap.png",
    width = 1800, height = 1200)
ComplexHeatmap::Heatmap(
  PCAresult$rotated[1:3] ,
  col = myColor2,
  column_split = 6,
  split = clusters,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(
      x, y, width = width, height = height,
      gp = gpar(col = "darkgrey", fill = NA, lwd = 1) # Add gray borders
    )
  }
) 
dev.off()


# Determine the optimal number of clusters
# Plot the dendrogram
plot(hc, main = "Hierarchical Clustering Dendrogram", sub = "", xlab = "", ylab = "Height", hang = -1, cex = 0.6)

# Cut the dendrogram into 3 clusters
clusters <- cutree(hc, k = 2)

# Add cluster assignments to the data frame
rot$cluster <- as.factor(clusters)





 rot = rot %>% rownames_to_column("IDsample") 
 methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    inner_join(rot) %>%
    pivot_wider(values_from = "methylation", names_from = "location" )  %>%
    ggplot(aes(x = GeneBody, y = TE, shape = Tissue,color = cluster)) + geom_point()+
    theme_light()+
    geom_text_repel(aes(label = IDsample), color = "black") -> fig3B.clustered
     

 PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary)  %>%
  inner_join(rot) %>%
  ggplot(aes(x = PC1, y = PC2, color =cluster, shape = Tissue,label =sampleID )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  ) +geom_text_repel(aes(label = sampleID), color = "black")  -> 
  
  PCA.plot.clustered
 PCA.plot.clustered
```


####  Non-Parametric Test (Kruskal-Wallis)



```{r }



```


#### Linear mixed model
```{r}


methylationData.3.QC.TE %>% distinct(TE_ID2) ->
  TE.unique

methylationData.3.QC.TE %>%
  select(TE_ID2,Sample,Tissue,start,percent) %>%
  ungroup() %>%
  group_by(TE_ID2,Tissue,start) %>%
    summarise(n = n(), methylation = mean(percent)) ->
  TE.summary.all
  
TE.summary.all %>%
  group_by(TE_ID2,Tissue) %>%
  summarise(n = n()) %>%
  ungroup %>%
  filter(n >=10) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n >=2) ->
  TE.filtered

TE.summary.all %>% filter(TE_ID2 %in% TE.filtered$TE_ID2) %>%
  ungroup() %>%
  distinct(TE_ID2,start) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n > 10) ->
  TE.filtered.nCpG.nObs
  



```



```{r, message = FALSE, warning = FALSE}



dist.list.Tissue = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Tissue,start,percent) ->
      mat
     model <- lmer(percent ~ Tissue + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Tissue[[TE]] = TEinfo
       
}

dist.list.Location = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Location,start,percent) ->
      mat
     model <- lmer(percent ~ Location + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Location[[TE]] = TEinfo
       
}

dist.list.Rep = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Replicate2,start,percent) ->
      mat
     model <- lmer(percent ~ Replicate2 + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Rep[[TE]] = TEinfo
       
}


TEspvalues.Tissue = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Tissue[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Tissue[[TE]][["Name"]],
  Test = "Tissue"
  ) ->
  TEspvalues.Tissue[[as.character(TE)]]
}

TEspvalues.Location = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Location[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Location[[TE]][["Name"]],
  Test = "Location"
  ) ->
  TEspvalues.Location[[as.character(TE)]]
}

TEspvalues.Rep = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Rep[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Rep[[TE]][["Name"]],
  Test = "Rep"
  ) ->
  TEspvalues.Rep[[as.character(TE)]]
}



bind_rows(TEspvalues.Tissue) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Tissue


bind_rows(TEspvalues.Location) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Location

bind_rows(TEspvalues.Rep) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Rep

lmm_results.TE = bind_rows(lmm_results.TE.Tissue,lmm_results.TE.Location,lmm_results.TE.Rep )


lmm_results.TE %>% ggplot(aes(x = pvalue, color = Test)) +
  geom_density( adjust = 1/10) +
  labs(
    title = "Density plot",
    x = "p-value",
    y = "Density",
    shape = "Class"
  ) + theme_light() ->
  TE.pvalue.density.plot

ggsave(here("figures","Figure3","3AA.LMM.distribution.plot.pdf"),
       plot = TE.pvalue.density.plot,width = 180, height = 180, units = "mm", dpi = 300)


lmm_results.TE %>% arrange(adj_pvalue) %>%
  filter(significant == "Yes") ->
  lmm_results.TE.significant
lmm_results.TE.significant %>% 
  group_by(Test) %>%
  summarise(numberOfSignificant = n()) ->
  lmm_results.TE.significant.summary



TEclass_stats <- TEclass_counts %>%
  rowwise() %>%
  mutate(
    p_value = fisher.test(matrix(c(No, Yes, 839 , 5968), nrow = 2))$p.value
  ) %>%
  ungroup() %>%
  mutate(ratio = Yes/No) %>%
  arrange(-ratio) %>%
  mutate(average =839/5968 ) %>%
  mutate(median = median(ratio))





```

###  Aalysis of  significant TEs  
```{r }    
lmm_results.TE.Tissue %>% filter(significant == "Yes") %>%
  arrange(adj_pvalue) ->
  lmm_results.TE.Tissue.significant

TE_tissue_DE = intersect(lmm_results.TE.Tissue.significant$TE_ID2,as.numeric( rownames(methylationData.3.QC.TE.wide))) 

methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]->

methylationData.3.QC.TE.sig.wide 





```
#### Cluster of TEs
##### Hierarchical clustering with correct number of clusters
```{r}


## Cluster analysis
dist_matrix <- dist(methylationData.3.QC.TE.sig.wide, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
library(NbClust)
nb <- NbClust(methylationData.3.QC.TE.sig.wide, distance="euclidean", min.nc=2, max.nc=10, method="ward.D2")


wss <- numeric(10)
for (k in 2:10) {
  clusters <- cutree(hc, k)
  wss[k] <- sum(sapply(unique(clusters), function(c) {
    pts <- methylationData.3.QC.TE.sig.wide[clusters == c, ]
    sum(scale(pts, scale=FALSE)^2)
  }))
}

plot(1:10, wss, type="b", xlab="k", ylab="Within-cluster SS")




nrOfClusters = nb$Best.nc[[1]]

nrOfClusters = 3
clusters <- cutree(hc, nrOfClusters)


clusters %>%  as.data.frame() %>%
  rownames_to_column(var = "TE_ID2") %>%
  rename(cluster = ".") ->
  clusters




```


#####  Visualisation of 3 clusters

```{r}


clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(cluster, Tissue) %>%
  summarise(mean = mean(percent), sd = sd(percent))->
  clusters.Tissue.summary


clusters.Tissue.summary %>% group_by(cluster) %>%
  summarise(overallMean = mean(mean)) %>%
  arrange(overallMean) %>%
  mutate(clusterOrder = as.factor(row_number()))->
  cluster.order
  



clusters %>%  
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  ungroup() %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n()) %>%
  arrange(cluster, -nrOfTEs)%>%
  group_by(cluster) %>%
  slice_head(n = 5)->
  clusters.count.summary


clusters.Tissue.summary %>%
  mutate(clusterName = "Low Methylation") %>%
  mutate(clusterName = replace(clusterName, 
                               grepl(1, cluster),
                               "Medium Methylation")) %>% 
  
  mutate(clusterName = replace(clusterName, 
                               grepl(3, cluster),
                               "High Methylation")) %>%
  mutate(clusterName = factor(clusterName, 
                              levels = c("Low Methylation",
                                         "Medium Methylation",
                                         "High Methylation"))) %>%
  inner_join(clusters.count.summary)->
clusters.summary

clusters.Tissue.summary %>%
  inner_join(clusters.count.summary) %>%
  inner_join(cluster.order)->
clusters.summary
  


clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(TE_ID2,Tissue,Sample,cluster ) %>%
  summarise(percent = mean(percent) )%>%
  inner_join(clusters.summary) %>%
  ggplot(aes(x = clusterOrder, y = percent, fill = Tissue)) +
  geom_boxplot() +
  theme_light() +
  labs(
    title = "Methylation levels per TE cluster",
    x = "TE Cluster",
    y = "Percent Methylation"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )+
  scale_fill_manual(values = colors) -> distributionPerClusterPlot
   
     

ha = HeatmapAnnotation(
    cluster = as.factor(clusters$cluster)
)

distributionPerClusterPlot

pdf("~/MarasmiusMethylationRNA/figures/Figure3/TE.methylation.clustered.heatmap.pdf",
    width = 8, height = 6)
ComplexHeatmap::Heatmap(t(methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]),
                        show_column_names = FALSE,
                        show_column_dend = TRUE, 
                        cluster_columns = hc,
                        column_split = nrOfClusters,
                        top_annotation = ha,
                        cluster_rows =TRUE,  split = 2
) 

dev.off()
distributionPerClusterPlot


clusters %>%   
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n())  %>%
  inner_join(cluster.order) %>%
  filter(TEclass %in% clusters.count.summary$TEclass) %>%
  mutate(TEfamily = TEclass) %>%
  inner_join(TE.classes) %>%
  ggplot(aes(x = clusterOrder, y = nrOfTEs, fill = Class)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_light() +
  labs(
    title = "TE class distribution per TE cluster",
    x = "TE Cluster",
    y = "Number of TEs"
  ) -> distribution


clusters %>%   
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n())  %>%
  inner_join(cluster.order) %>%
  filter(TEclass %in% clusters.count.summary$TEclass) %>%
  mutate(TEfamily = TEclass) %>%
  inner_join(TE.classes) %>%
  ggplot(aes(x = clusterOrder, y = nrOfTEs, fill = TEfamily)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_light() +
  labs(
    title = "TE class distribution per TE cluster",
    x = "TE Cluster",
    y = "Number of TEs"
  ) -> distribution.TEfamily



cluster.order

clusters%>%   
  inner_join(clusters.count.summary) %>%   
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n())  %>%
  inner_join(cluster.order) %>%
  ggplot(aes(x = clusterOrder, y = nrOfTEs)) +
  geom_bar(stat = "identity") +
  theme_light() +
  labs(
    title = "TE class distribution per TE cluster",
    x = "TE Cluster",
    y = "Number of TEs"
  ) -> distribution2





distributionPerClusterPlot+distribution2+
  distribution+ distribution.TEfamily+ plot_layout(ncol = 2, guides = "collect") ->
  MLM.Methylation.distribution
  
ggsave(here("figures","Figure3","LMM.classes.plot.pdf"),
       plot = MLM.Methylation.distribution,width = 180, height = 180, units = "mm", dpi = 300)


```

