---
title: "Methylation pattern"
output: html_document
date: "2024-09-11"
---

## setup
```{r setup, include=FALSE}

#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#library(ropls)

#install.packages("pls")
#library(pls)
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(GenomicFeatures)
library(tidyverse)
library(here)


#  BiocManager::install('PCAtools')
library(PCAtools)
library(patchwork)
library(gridExtra)


select = dplyr::select


types = c("Stipe","Lamella","Spore","control")
colors = c("#96c176ff","#cf9ddbff","#4472C4ff","#000000FF")
types = c("Stipe","Lamella","Spore","control","stipe","lamella","spore")
colors = c("#96c176ff","#cf9ddbff","#4472C4ff","#000000FF","#96c176ff","#cf9ddbff","#4472C4ff")
names(colors) = types
  

annotationColors = c("mRNA" = "#4472C4","TE" = "#C00000","Both" = "#9D475B","intergenic" = "#AFABAB")



methylationLevelColors = my_colors <- c(
  "Nothing"   = "#4575b4",
  "Low"  = "#91bfdb",
  "Medium"  = "#e0f3f8",
  "High"  = "#fee090",
  "All" = "#d73027"
)



```




## Annotation


## Samples 
```{r}

select = dplyr::select

samples.ring.methyl.df = read_csv(file = here("data","meDNA","meDNA.metaData.csv"))


samples.ring.methyl.df %>% arrange(Tissue,Location,Replicate) -> 
  samples.ring.methyl.df


samples.ring.methyl.df %>% group_by(Tissue,Location) %>% summarise(n = n()) %>%
  pivot_wider(names_from = "Location", values_from = "n")->
  samples.ring.methyl.df.summary



# make the directory if it does not exist
dir.create(here("results","meDNA"), showWarnings = FALSE, recursive = TRUE)

# Open the PDF device
pdf(here("results","meDNA","meDNA.samples.summary"), width = 8, height = 11)
# Draw the table
grid.table(samples.ring.methyl.df.summary)
# Close the device
dev.off()

```

## Data 
Data from Nextflow transformed into one table in LoadMethylationData.Rmd
```{r}
methylationDataFile = "~/MarasmiusMethylationRNA/data/methylationData.samples.csv"
methylationDataFile.QC = "~/MarasmiusMethylationRNA/data/methylationData.samples.QC.csv.gz"

if(!file.exists(methylationDataFile.QC)){
  methylationData.3 = read_csv(methylationDataFile)
  # Remove sites with no sample data
  methylationData.3 %>% filter(!is.na(Sample))->methylationData.3
}
```


### TE reference
```{r}
gff3Colnames =c("seqid", "source",
                "type","start",
                "end", "score",
                "strand","phase",
                "attributes")


  TE.classes = read_csv(file = here("annotation","Repeat.classes.csv")) %>%
    rename(TEfamily = "Name") %>%
    rename(Notes = "Superfamily / Notes")

  
  TEFile <- here("annotation","Maror2.repeatmasker.rmlib2.3.gff")

  TEdf =    read_tsv(file = TEFile, col_names = gff3Colnames,comment = "#") %>% separate(attributes, into = c("ID","target"), sep = ";") %>%
    mutate(ID = as.numeric(gsub("ID=","",ID))) %>%
    separate(target, into = c("Target","motif",
                              "motifStart","motifStop"),
             sep = " ") %>%
    mutate(motif = gsub("\"","",motif))%>%
    mutate(motif = gsub("Motif:","",motif))  %>%
    select(seqid,start,end,score,
           strand, ID,motif,
           motifStart,motifStop) %>%
    rename(TEID = "ID") %>%
    mutate(TE_ID2 = 1:n()) %>%
    mutate(motifStart = as.numeric(gsub(" ","",motifStart))) %>%
    mutate(motifStop = as.numeric(gsub(" ","",motifStop))) %>%
    dplyr::rename(TEclass = "motif") %>%
    mutate(TEclass = gsub(" ","",TEclass)) %>%
    mutate(TEclass = gsub("Maror","",TEclass)) 

 

  TEs = makeGRangesFromDataFrame(TEdf,
                                 seqnames.field = "seqid", 
                                 start.field = "start", 
                                 end.field = "end",
                                 strand.field = "strand",
                                 keep.extra.columns = TRUE)
  

TEdf%>%
  separate(TEclass, into = c("TEfamily","nr"), sep = "-",remove = FALSE) %>%
  inner_join(TE.classes)->
  TEdf

TEdf %>%
  group_by(TEfamily) %>% summarise(n = n(), min = min(motifStart), max = max(motifStop)) %>%
  mutate(TElength = max - min) %>%
  filter(TElength > 1000) %>%
  inner_join(TEdf) %>%
  mutate(fraction =(motifStop-motifStart) / TElength)->
  TEdf.long

TEdf.long  %>% separate(TEclass, into = c("TEfamily","nr"), sep = "-",remove = FALSE) ->
  TEdf.long

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(grepl("DNA",Class)) %>%
  ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(Class~TEfamily, scales = "free_x", space = "free") -> 
  TE.distribution.ClassII.plot

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(grepl("Retro",Class)) %>%
    ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(Class~TEfamily, scales = "free_x", space = "free") -> 
 TE.distribution.ClassI.plot

TEdf.long %>% 
  inner_join(TE.classes) %>% 
  filter(!grepl("DNA",Class)) %>%
  filter(!grepl("Retro",Class)) %>%
  ggplot(aes(x = nr, y = fraction)) + 
  geom_boxplot() +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ 
  facet_grid(.~TEfamily, scales = "free_x", space = "free") -> 
 TE.distribution.Other.plot


TE.distribution.ClassI.plot+
  TE.distribution.ClassII.plot+
  TE.distribution.Other.plot +
  plot_annotation(tag_levels = 'A')+plot_layout(ncol = 1) ->
  TE.distribution.plots

ggsave(here("results","annotation","TE.distribution.plots.pdf"),
       TE.distribution.plots,
       width = 10, height = 15, units = "in", dpi = 300)

  
TEdf.long %>% arrange(TEclass, -fraction) %>% 
  filter(fraction > 0.25)

```


### Gene reference 
```{r}

gtfFile = here("annotation","GCF_018924745.1_UU_Maror_2_genomic.gff")	
txdb <- makeTxDbFromGFF(file=gtfFile, dataSource="ensemblgenomes",format = "gff3")


gff3Colnames =c("seqid", "source",
                "type","start",
                "end", "score",
                "strand","phase",
                "attributes")

  annotation.df =    read_tsv(file = gtfFile, col_names = gff3Colnames,comment = "#") 


 annotation.df %>% filter(type =="mRNA") %>%
   group_by(seqid,start,end, strand) %>%
   slice_head(n = 1) %>%
   ungroup() %>%
   separate_longer_delim(attributes, delim = ";") ->
   test
 
  test %>% select(seqid,start,end,strand,attributes) %>%
   separate(attributes, into = c("key","value"), sep = "=") %>%
   filter(key %in% c("Name","locus_tag","product")) %>%
   pivot_wider(values_from = "value",names_from = "key") -> 
   annotation.mRNA.df

 annotation.mRNA.df %>% group_by(product) %>% summarise(n = n())
 
annotation.mRNA.df %>% select(seqid,start,end, strand,Name,locus_tag,product) %>%
  mutate(tx_id = 1:n()) -> 
  annotation.mRNA.df

mRNAs = makeGRangesFromDataFrame(annotation.mRNA.df,
                                 seqnames.field = "seqid", 
                                 start.field = "start", 
                                 end.field = "end",
                                 strand.field = "strand",
                                 keep.extra.columns = TRUE)
  

```

Run only once for getting relationship between protein ID (XP) and gene ID (XM)
```{r, inlcude false  }

gene2ProteinFile = here("annotation","gene2protein_relationship.csv")
if(file.exists(gene2ProteinFile)){
  relationship_df = read_csv(gene2ProteinFile)
# Load the GFF file
gff_df <- read_tsv(gtfFile, comment = "#", col_names = FALSE)


filtered_gff <- gff_df %>%
  filter(grepl("Parent=rna-XM_", X9) | grepl("ID=cds-XP_", X9))

# Use mutate with regex replacements to extract and clean XM_ID and XP_ID
relationship_df <- filtered_gff %>%
  mutate(
    XM_ID = sub(".*Parent=rna-(XM_[^;]+).*", "\\1", X9),
    XP_ID = sub(".*ID=cds-(XP_[^;]+).*", "\\1", X9)
  ) %>%
  # Filter out rows where either XM_ID or XP_ID are missing or invalid
  filter(grepl("XM_", XM_ID) & grepl("XP_", XP_ID)) %>%
  # Select only the XM_ID and XP_ID columns and keep distinct pairs
  select(XM_ID, XP_ID) %>%
  distinct()

write_csv(relationship_df,gene2ProteinFile)
}else{
  relationship_df = read_csv(gene2ProteinFile)
}

relationship_df %>% mutate(Name = XM_ID) -> relationship_df
annotation.mRNA.df = inner_join(annotation.mRNA.df,relationship_df , by = "Name")
```

### Distance between RNA and TEs
```{r}
# Find the nearest TE for each mRNA
# query = mRNAs, subject = TEs
nearest_hits <- distanceToNearest(mRNAs, TEs, ignore.strand=TRUE)
# Extract results into a readable format
# This creates a data frame with mRNA index, TE index, and the distance
results <- as.data.frame(nearest_hits)

# Optional: Merge metadata to see which mRNA is near which TE
results$locus_tag <- mRNAs$locus_tag[results$queryHits]
results$TE_ID2   <- TEs$TE_ID2[results$subjectHits] # Replace 'name' with your TE column name

results %>%
  select(locus_tag, TE_ID2, distance) ->
  GeneBody_TE_distance

```





## Initial data analysis
```{r}
if(!file.exists(methylationDataFile.QC)){

methylationData.3 %>% group_by(Location,Tissue) %>%
  mutate(Replicate2 = dense_rank(Replicate))->
  methylationData.3



methylationData.3 %>% group_by(Location,Replicate2,Replicate,Tissue) %>%
  summarise(n = n(), mean = mean(depth), median = median(depth)) -> 
  methylationData.3.summary

methylationData.3.summary %>% 
  mutate(Location = factor(Location, levels = c("west","northwest","north","northeast")))%>%
  mutate(Tissue = factor(Tissue, levels = c("stipe","lamella","spore"))) %>%
  mutate(IDsample = paste(Tissue,Location,Replicate2,sep = "_")) %>%
  arrange(Location,Replicate2,Tissue) %>%
  mutate(IDsample = factor(IDsample, levels = IDsample)) ->
  methylationData.3.summary


methylationData.3.summary %>%
  ggplot(aes(x = IDsample, y = mean, fill = Tissue),color = "grey") +
  geom_hline(yintercept = 10, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5)) +  
  facet_grid(.~Location, space = "free", scales = "free_x") ->
  methylationData.3.summary.plot

methylationData.3.summary %>%
  dplyr::select(Location, Replicate2,Tissue,mean) %>%
  arrange(Replicate2,Tissue) %>%
  ungroup() %>%
  dplyr::select(-Replicate) %>%
  pivot_wider(values_from = "mean", names_from = "Location")%>%
  dplyr::rename(replicate = "Replicate2") ->
  methylationData.3.summary.wide
  

my_table <- tableGrob(methylationData.3.summary.wide)

# Wrap the table for patchwork
table_plot <- wrap_elements(my_table)


write_csv(methylationData.3.summary.wide,"~/MarasmiusMethylationRNA/data/methylationData.samples.summary.csv")
methylationData.3.summary.plot/table_plot
ggsave("~/MarasmiusMethylationRNA/suppFigures/methylationData.samples.summary.pdf",width = 10, height = 8, units = "in", dpi = 300)

}
```


### QC intial
Done only once!
```{r}
if(!file.exists(methylationDataFile.QC)){
methylationData.3.summary %>%
  ggplot(aes(x = IDsample, y = mean, fill = Tissue)) +
  geom_bar(stat = "identity", color = "grey") +
  geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
  scale_fill_manual(values = colors) + 
  facet_grid(. ~ Location, space = "free", scales = "free_x") +
  labs(
    title = "Mean Methylation Levels per Sample",
    x = "Sample ID",
    y = "Average reads depth",
    fill = "Tissue Type"
  ) +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.5),
    plot.title = element_text(hjust = 0.5)
  ) ->
  methylationData.3.summary.plot

methylationData.3.summary %>% filter(mean > 10) -> methylationData.3.summary.qc


methylationData.3 %>% 
  group_by(seqid,start) %>%
  summarise(depth = mean(depth), percent = mean(percent)) %>%
  ungroup() ->
  methylationData.3.summary.sites

methylationData.3 %>% inner_join(methylationData.3.summary.qc) ->
  methylationData.3.QC



methylationData.3.QC %>% 
  ggplot(aes(x = depth)) + 
  geom_density(fill = "steelblue", alpha = 0.6, adjust = 5) +  # smoother line
  geom_vline(xintercept = 10, linetype = "dashed", color = "red") +  # threshold
  annotate("text", x = 2, y = Inf, label = "Depth < 10 removed", 
           vjust = 2, hjust = 0, color = "red", size = 3.5) +
  scale_x_log10() +
  coord_cartesian(xlim = c(1, 100)) +
  labs(
    title = "Distribution of Methylation Depth (log10)",
    x = "Depth (log10 scale)",
    y = "Density"
  ) +
  theme_minimal() ->QC.10filter.plot

methylationData.3.QC %>% filter(depth > 10 ) -> methylationData.3.QC



## Removing sites that are all 100% methylated or not methylated at all
methylationData.3.QC %>% group_by(seqid,start) %>%
  summarise(n = n(), percent = mean(percent)) -> 
methylationData.3.QC.location


methylationData.3.QC.location %>%
  ggplot(aes(x = percent)) +
  geom_density(fill = "steelblue", alpha = 0.6, adjust = 0.02) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 100, linetype = "dashed", color = "red") +
  annotate("text", x = 0.02, y = Inf, label = "0% average methylation removed",
           vjust = 2, hjust = 0, color = "red", size = 3.5) +
  annotate("text", x = 100, y = Inf, label = "100% average methylation removed",
           vjust = 2, hjust = 1, color = "red", size = 3.5) +
  labs(
    title = "Distribution of Mean Methylation Across Samples",
    x = "Mean Methylation (0 to 100 percent)",
    y = "Density"
  ) +
  theme_minimal() -> methylationData.3.QC.location.plot
  

methylationData.3.QC.location %>% filter( percent > 0 & percent < 100) %>%
  select(seqid,start) %>% inner_join(methylationData.3.QC) ->
  methylationData.3.QC


methylationData.3.QC.location %>% filter( percent > 0 & percent < 100)->
  methylationData.3.QC.location
## Removing sites with depth less than 10 reads

methylationData.3.QC.location %>%
  ggplot(aes(x = n)) +
  geom_histogram(fill = "steelblue", alpha = 0.6,binwidth = 1) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "red") +
  annotate("text", x = 10, y = Inf, label = "sites found in less than 10 samples removed",
           vjust = 2, hjust = 1, color = "red", size = 3.5) +
  labs(
    title = "Samples across sites",
    x = "Number of samples with\n coverage at a site (n)",
    y = "Number of sites"
  ) +
  theme_minimal() -> nrOfSamples.plot2

methylationData.3.QC.location %>%ungroup() %>% distinct(n) 

  methylationData.3.QC.location %>% filter(n > 10) ->
    methylationData.3.QC.location
    
    
  
methylationData.3.QC.location%>%
  select(seqid,start) %>%
  inner_join(methylationData.3.QC) ->
  methylationData.3.QC

layout  <- "AB
CC
DD"

methylationData.3.summary.plot + 
  QC.10filter.plot +
  methylationData.3.QC.location.plot+
  nrOfSamples.plot2 + 
  plot_layout(design = layout)+
  plot_annotation(tag_levels = 'A')->
  QC.plots
ggsave("~/MarasmiusMethylationRNA//suppFigures/QC.methylation.pdf",
       QC.plots, # your patchwork object
      width = 21,
      height = 24,
  units = "cm"
       )

ggsave("~/MarasmiusMethylationRNA/suppFigures/QC.methylation.png",
       dpi = 300,
       QC.plots, # your patchwork object
      width = 21,
        
      height = 24,
  units = "cm"
       )



write_csv(methylationData.3.QC,methylationDataFile.QC)
}else{
  methylationData.3.QC = read_csv(methylationDataFile.QC)
}

methylationData.3.QC %>% 
  mutate(location = gsub("mRNA","GeneBody",location )) %>%
    mutate(location = gsub("intergenic","Intergenic",location ))->
  methylationData.3.QC

```




### Remove  sites annotated as both

```{r}
methylationData.3.QC %>% filter(location != "Both") ->
  methylationData.3.QC


methylationData.3.QC %>% distinct(location)
```

# Figure2

Change to Intergenic (Capital) (DONE)
Change to fraction instead of percent (DONE)
Change to Gene body instead of GeneBody (Not done make in ai)
Make a statistical test for the differences in methyalation level.



```{r}


methylationData.3.QC %>% 
  group_by(location,Tissue,Location,Replicate2) %>% 
  summarise(n = n(), depth = mean(depth), methylation = mean(percent)) ->
  methylationData.3.QC.annotation.distribution


 

methylationData.3.QC.annotation.distribution %>%
  filter(location != "Both") %>%
  select(Tissue,Location,Replicate2,methylation,location) %>%
  mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
  mutate(fraction = methylation /100)%>%
  ggplot(aes(x = Tissue, y = fraction, fill = Tissue)) + # Change x to Tissue
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.2) + 
#  geom_text(aes(label =IDsample ))+ # Add alpha for better visualization if many points
  theme_light() +
  scale_fill_manual(values = colors) +
  facet_wrap(.~location, scales = "free_y") -> fig2C

ggsave("~/MarasmiusMethylationRNA/figures/Figure2/fig2C.methylationData.samples.annotation.distribution.1.pdf",plot = fig2C,
       width = 8, height = 5, units = "in", dpi = 300)



methylationData.3.QC.annotation.distribution %>%
  filter(location != "Both") %>%
  select(Tissue,Location,Replicate2,methylation,location) %>%
  mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
  mutate(fraction = methylation /100)%>%
  ggplot(aes(x = location, y = fraction, fill = location)) + # Change x to Tissue
  geom_boxplot(outliers = FALSE) +
  geom_jitter(width = 0.2) + 
#  geom_text(aes(label =IDsample ))+ # Add alpha for better visualization if many points
  theme_light() +
  ylim(c(0,1))+
  scale_fill_manual(values = colors)  -> fig2A

ggsave("~/MarasmiusMethylationRNA/figures/Figure2/fig2A.methylationData.samples.annotation.distribution.1.pdf",plot = fig2A,
       width = 4, height = 5, units = "in", dpi = 300)


```


```{r}

methylationData.3.QC %>% 
  group_by(Tissue,Location,Replicate) %>% 
  summarise(n = n(), depth = mean(depth), methylation = mean(percent)) ->
  methylationData.3.QC.distribution


methylationData.3.QC.distribution %>% 
  mutate(IDsample = paste(Tissue,Location,Replicate,sep = "_")) %>%
  ggplot(aes(x = IDsample, y = depth, fill = Tissue),color = "grey") +
  geom_hline(yintercept = 10, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5)) 
  
  

methylationData.3.QC.annotation.distribution %>% 
  dplyr::rename(annotation = "location") %>%
  mutate(IDsample = paste(Tissue,Location,Replicate2,sep = "_")) %>%
  ggplot(aes(x = IDsample, y = depth, fill = annotation),color = "grey") +
  geom_bar(stat = "identity",position = "fill") +
  scale_fill_manual(values = annotationColors) + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5)) ->
  distribution.depth

methylationData.3.QC.annotation.distribution %>% 
  dplyr::rename(annotation = "location") %>%
  mutate(IDsample = paste(Tissue,Location,Replicate2,sep = "_")) %>%
  ggplot(aes(x = IDsample, y = methylation, fill = annotation),color = "grey") +
  geom_bar(stat = "identity",position = "dodge") +
  scale_fill_manual(values = annotationColors) + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5)) 




methylationData.3.QC %>% group_by(Tissue,Location,Replicate) %>%
  summarise(sum = sum(n())) %>% 
  inner_join(methylationData.3.QC.annotation.distribution) %>%
  mutate(fraction = n / sum) %>%
  mutate(location = factor(location, levels = c("mRNA","Both","TE","intergenic"))) %>%
  dplyr::rename(annotation = "location") %>%
  ggplot(aes(x = Tissue, y = n, fill = annotation)) +geom_bar(stat = "identity", color = "black") +
  facet_wrap("Location")+ 
  scale_fill_manual(values = annotationColors) 
#ggsave("../../results/meDNA/locationDistribution.pdf")

 

  methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    pivot_wider(values_from = "methylation", names_from = "location" )  %>%
    ggplot(aes(x = GeneBody, y = TE, color = Tissue)) + geom_point()+
    theme_light()+
    geom_text_repel(aes(label = IDsample), color = "black")  +
    scale_color_manual(values = colors) -> sampleDistribution
#ggsave("../../results/meDNA/sampleDistribution.pdf")
  
methylationData.3.QC.annotation.distribution %>% 
  select(Tissue,Location,Replicate2,methylation,location) %>%
  mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
  filter(IDsample %in%
           c("northeast_1_stipe","north_3_stipe","north_3_spore")) %>%
    mutate(fraction = methylation /100) %>%
    select(-methylation) %>%
    pivot_wider(values_from = "fraction", names_from = "location" ) ->outliers  
  

  methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    mutate(fraction = methylation /100) %>%
    select(-methylation) %>%
    pivot_wider(values_from = "fraction", names_from = "location" )  %>%
    ggplot(aes(x = GeneBody, y = TE, color = Tissue)) + geom_point()+
    theme_light()+
    geom_text_repel(data = outliers,aes(label = IDsample), color = "black")  +
    scale_color_manual(values = colors) -> 
    fig2B
  
  
ggsave("~/MarasmiusMethylationRNA/figures/Figure2/Fig2B.TE.GeneBody.scatterPlot.pdf",plot = fig2B,
       width = 6, height = 5, units = "in", dpi = 300)


methylationData.3.QC %>%
  mutate(methylated = depth*percent /100) %>%
  mutate(notMethylated = depth - methylated) %>%
  group_by(Sample) %>%
  summarise(methylated = sum(methylated), notMethylated = sum(notMethylated)) %>%
  mutate(fraction = methylated /(notMethylated+methylated))->
  sum

sum %>%
  inner_join(samples.ring.methyl.df) %>%
  ggplot(aes(x = Tissue, y = fraction,fill = Tissue,)) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(width = 0.1,mapping = aes(shape =Location)) + 
  theme_light()+
  scale_fill_manual(values = colors) ->
  averageMethylation


methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    mutate(fraction = methylation /100) %>%
  group_by(location) %>%
  summarise(
    p_value = tryCatch({
      kruskal.test(fraction ~ Tissue, data = cur_data())$p.value
    }, error = function(e) NA),
  )%>% 
  mutate(comparison = "Location") %>%
  mutate(adj_pvalue = p.adjust(p_value, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.05), "Yes")) ->
  kruskal_results.Methylation


library(rstatix)
library(purrr)

results <- methylationData.3.QC.annotation.distribution %>% 
  select(Tissue, Location, Replicate2, methylation, location) %>%
  mutate(
    IDsample = paste(Location, Replicate2, Tissue, sep = "_"),
    fraction = methylation / 100
  ) %>%
  group_by(location) %>%
  group_modify(~{
    
    df <- .x  # current location
    
    ## --- Kruskal–Wallis test ---
    kw <- tryCatch(
      kruskal.test(fraction ~ Tissue, data = df),
      error = function(e) NULL
    )
    
    kw_p <- if (!is.null(kw)) kw$p.value else NA
    
    ## --- Dunn test (pairwise comparisons) ---
    dunn <- df %>%
      dunn_test(fraction ~ Tissue, p.adjust.method = "BH") %>%
      mutate(
        kw_p_value = kw_p
      )
    
    dunn
  }) %>%
  ungroup() %>%
  
  ## Adjust the Kw p-values across all locations
  group_by(location) %>%
  mutate(
    kw_p_adj = p.adjust(unique(kw_p_value), method = "BH")
  ) %>%
  ungroup() %>%
  
  ## Add significance flags
  mutate(
    kw_significant = kw_p_adj < 0.05,
    pairwise_significant = p.adj < 0.05
  )

write_csv(results, "~/MarasmiusMethylationRNA/figures/Figure2/fig2C.methylationData.samples.annotation.distribution.kruskal.dunn.test.results.csv")

```




```{r}


methylationData.3.QC %>% 
  group_by(Tissue, location) %>%
  summarise(mean = mean(percent))
  
methylationData.3.QC %>% 
  ggplot(aes(x = percent,color = Tissue )) +
  geom_density()+
  geom_vline(xintercept = 90, linetype = "dashed", color = "red")+
  geom_vline(xintercept = 10, linetype = "dashed", color = "blue")+
  scale_color_manual(values = colors)  +
  theme_minimal() +
  labs(
    fill = "Methylation Level", 
    x = "Precent methylated per site", 
    y = "Density"
  )->methylation.density.plot
  



methylationData.3.QC %>% 
  mutate(methylationLevel = "High") %>%
  mutate(methylationLevel = replace(methylationLevel,
                                    percent <90,
                                    "Average" )) %>%
  mutate(methylationLevel = replace(methylationLevel,
                                    percent <10,
                                    "Low" )) ->
  methylationData.3.QC






methylationData.3.QC %>% 
  group_by(Location,location, Tissue,methylationLevel) %>%
  summarise(n = n()) -> methylationData.3.QC.summary

methylationData.3.QC.summary %>%
  group_by(Location,location, Tissue) %>%
  summarise(Total = sum(n)) %>% 
  inner_join(methylationData.3.QC.summary) %>%
  mutate(fraction = n /Total) -> methylationData.3.QC.summary

methylationData.3.QC <- methylationData.3.QC %>%
  mutate(methylationLevel = factor(methylationLevel, levels = c("High", "Average", "Low")))


ggplot(methylationData.3.QC, aes(x = Tissue, fill = methylationLevel)) +
  geom_bar(position = "fill") +
  facet_wrap(facets = vars(location)) + # Ensure 'location' is used properly
  scale_fill_manual(values = c("Low" = "blue", "Average" = "grey", "High" = "red")) +
  theme_minimal() +
  labs(
    fill = "Methylation Level", 
    x = "Tissue", 
    y = "Count"
  )-> methylationDistribution
ggsave("../../results/meDNA/MethylationDistribution.pdf")


layout= c("AB
          CC
          CC")
#methylation.density.plot+ methylationDistribution+
#    sampleDistribution + plot_layout(design = layout, guides = "collect") +
#  plot_annotation(tag_levels = 'A')-> figure2
#ggsave("~/MarasmiusMethylationRNA/suppFigures/Figure2.methylation.pdf",width = 10, height = 8, units = "in", dpi = 300,plot = figure2)

#ggsave("~/MarasmiusMethylationRNA/figures/Figure2.methylation.annotation.png",
#       plot = figure2, width = 180,
#       height = 180, units = "mm", 
#       dpi = 300,plot = figure2)








```
#Figure 3

Add an pvalue line to the distribution
Add a heatmap with only the significant ones

Check that the mRNA and TE analysis are similar in terms of significant sites

###  TE analysis

```{r }
TEdf %>% 
  mutate(motifStart = as.numeric(motifStart)) %>%
  mutate(motifStop = as.numeric(motifStop)) %>%
  separate(TEclass, into=c("TEclass","nr"), sep = "-") ->
  TEdf
TEdf %>%
  group_by(TEclass) %>%
  summarise(motifStart = min(motifStart), motifStop = max(motifStop)) %>%
  arrange(-motifStop)  %>%
  filter(motifStop>500) %>%
  select(TEclass) %>%
  inner_join(TEdf)->
  TEdf.long

TEdf.long %>% select(TEclass,seqid,TE_ID2) -> 
  TE.info

TE.info 
  


methylationData.3.QC %>% 
  select(-TEclass) %>%
  filter(location == "TE") %>%
  inner_join(TE.info) ->
  methylationData.3.QC.TE
```


####PCA analysis

```{r }
methylationData.3.QC.TE %>%
  group_by(TE_ID2,Location,Replicate2,Tissue) %>%
  summarise( percent = mean(percent)) ->
  methylationData.3.QC.TE.summary


methylationData.3.QC.TE.summary %>%
  group_by(TE_ID2,Tissue) %>%
  summarise(nrOfExamples = n()) -> methylationData.3.QC.TE.summary.2

methylationData.3.QC.TE.summary.2 %>%
  group_by(Tissue) %>%
  summarise(n = max(nrOfExamples)) %>%
  inner_join(methylationData.3.QC.TE.summary.2) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n == 3)  %>%
  select(TE_ID2)->
  TE.covered


TE.covered
  

methylationData.3.QC.TE.summary %>%
  inner_join(TE.covered) %>%
  pivot_wider(values_from = "percent",names_from = c("Location","Replicate2","Tissue")) %>%
  column_to_rownames(var = "TE_ID2")%>%
  drop_na()->                     # Removes any row with at least one NA->
  methylationData.3.QC.TE.wide


## PCA analysis
PCAresult = PCAtools::pca(methylationData.3.QC.TE.wide,center = TRUE)
rot = PCAresult$rotated[,1:6]

## Cluster analysis
dist_matrix <- dist(rot, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
library(NbClust)
nb <- NbClust(rot, distance="euclidean", min.nc=2, max.nc=10, method="ward.D2")
nb$Best.nc


wss <- numeric(10)
for (k in 2:10) {
  clusters <- cutree(hc, k)
  wss[k] <- sum(sapply(unique(clusters), function(c) {
    pts <- rot[clusters == c, ]
    sum(scale(pts, scale=FALSE)^2)
  }))
}

plot(1:10, wss, type="b", xlab="k", ylab="Within-cluster SS")


## Variance
PCA.stats = data.frame(variance = PCAresult$variance, 
                       PC = factor(names(PCAresult$variance), levels = names(PCAresult$variance)))
ggplot(PCA.stats, aes(x = PC, y = variance) )+
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(
    title = "PC variance",
    x = "PC",
    y = "Variance"
  )+theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  ) -> variation


ggsave("~/MarasmiusMethylationRNA/figures/figure3/TE.PCA.variance.plot.pdf",
       plot = variation, width = 10, height = 8, units = "in")


methylationData.3.QC.TE %>%
  group_by(Location,Replicate2,Tissue) %>%
  summarise(n = n(),mean = mean(mean)) %>%
  mutate(sampleID = paste(Location,Replicate2,Tissue, sep = "_")) ->
  sampleInfo.TE.summary

sampleInfo.TE.summary

PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary) %>%
  ggplot(aes(x = PC1, y = PC2, color =Tissue, shape = Location )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  scale_color_manual(values = colors) +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  )-> PCA.plot


  
PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary)  %>%
  ggplot(aes(x = PC1, y = PC2, color =Tissue, shape = Location,label =sampleID )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  scale_color_manual(values = colors) +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  ) + geom_text_repel() -> PCA.plot.labels



PCA.plot.labels


breaks1 <- seq(from = -100, to = 300, length.out = 4)
library(RColorBrewer)
colors1 = colorRampPalette(rev(brewer.pal(n = 7, name =
  "RdYlBu")))(4)
myColor2 <-circlize::colorRamp2(breaks = breaks1,colors =colors1)


library(ComplexHeatmap)

pdf("~/MarasmiusMethylationRNA/figures/Figure3/TE.methylation.PCA.heatmap.pdf",
    width = 8, height = 6)
ComplexHeatmap::Heatmap(
  PCAresult$rotated[1:3] ,
  col = myColor2,
  column_split = 6,
  split = clusters,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(
      x, y, width = width, height = height,
      gp = gpar(col = "darkgrey", fill = NA, lwd = 1) # Add gray borders
    )
  }
) 
dev.off()


png("~/MarasmiusMethylationRNA/figures/Figure3/TE.methylation.PCA.heatmap.png",
    width = 1800, height = 1200)
ComplexHeatmap::Heatmap(
  PCAresult$rotated[1:3] ,
  col = myColor2,
  column_split = 6,
  split = clusters,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(
      x, y, width = width, height = height,
      gp = gpar(col = "darkgrey", fill = NA, lwd = 1) # Add gray borders
    )
  }
) 
dev.off()


# Determine the optimal number of clusters
# Plot the dendrogram
plot(hc, main = "Hierarchical Clustering Dendrogram", sub = "", xlab = "", ylab = "Height", hang = -1, cex = 0.6)

# Cut the dendrogram into 3 clusters
clusters <- cutree(hc, k = 2)

# Add cluster assignments to the data frame
rot$cluster <- as.factor(clusters)





 rot = rot %>% rownames_to_column("IDsample") 
 methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    inner_join(rot) %>%
    pivot_wider(values_from = "methylation", names_from = "location" )  %>%
    ggplot(aes(x = GeneBody, y = TE, shape = Tissue,color = cluster)) + geom_point()+
    theme_light()+
    geom_text_repel(aes(label = IDsample), color = "black") -> fig3B.clustered
     

 PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary)  %>%
  inner_join(rot) %>%
  ggplot(aes(x = PC1, y = PC2, color =cluster, shape = Tissue,label =sampleID )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  ) +geom_text_repel(aes(label = sampleID), color = "black")  -> 
  
  PCA.plot.clustered
 PCA.plot.clustered
```


####  Non-Parametric Test (Kruskal-Wallis)



```{r }
#methylationData.3.QC.TE %>%
  #ggplot(aes(x = Tissue, fill = methylationLevel ))+ geom_bar(position = "fill") +  # Replace with your desired geom
#  theme(axis.text.x = element_text(angle = 70, hjust = 1)) + facet_wrap("TEclass")



methylationData.3.QC.TE %>%
  group_by(TE_ID2) %>%
  summarise(
    p_value = tryCatch({
      kruskal.test(percent ~ Location, data = cur_data())$p.value
    }, error = function(e) NA),
  ) %>%
  mutate(comparison = "Location") %>%
  mutate(adj_pvalue = p.adjust(p_value, method = "BH")) %>%
  inner_join(TE.info) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.001), "Yes")) ->
  kruskal_results.TE.location

kruskal_results.TE.location %>% group_by(significant)  %>% summarise(n = n())

methylationData.3.QC.TE %>%
  group_by(TE_ID2) %>%
  summarise(
    p_value = tryCatch({
      kruskal.test(percent ~ Replicate, data = cur_data())$p.value
    }, error = function(e) NA),
  ) %>%
  mutate(comparison = "Replicate") %>%
  mutate(adj_pvalue = p.adjust(p_value, method = "BH")) %>%
  inner_join(TE.info) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  kruskal_results.TE.replicate


kruskal_results.TE.replicate %>% group_by(significant)  %>% summarise(n = n())

methylationData.3.QC.TE %>%
  group_by(TE_ID2) %>%
  summarise(
    p_value = tryCatch({
      kruskal.test(percent ~ Tissue, data = cur_data())$p.value
    }, error = function(e) NA),n = n()
  ) %>%
  mutate(comparison = "Tissue") %>%
  mutate(adj_pvalue = p.adjust(p_value, method = "BH")) %>%
  inner_join(TE.info) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  kruskal_results.TE



bind_rows(kruskal_results.TE,kruskal_results.TE.replicate,kruskal_results.TE.location)   %>%
  ggplot(aes(x = p_value, color = comparison)) +
  geom_density( adjust = 1/1) +
  labs(
    title = "Density plot",
    x = "p-value",
    y = "Density",
    shape = "Class"
  ) + theme_light() ->
  TE.pvalue.density.plot
ggsave("~/MarasmiusMethylationRNA/figures/Figure3/3A.kruskal.distribution.plot.pdf",plot = TE.pvalue.density.plot,width = 180, height = 180, units = "mm", dpi = 300)



# Filter significant results
kruskal_results.TE %>%
  filter(!is.na(p_value) & adj_pvalue < (0.001))  %>%
  arrange(adj_pvalue)->
  kruskal_results.TE.significant

bind_rows(kruskal_results.TE,kruskal_results.TE.replicate,kruskal_results.TE.location) %>%
  filter(!is.na(p_value) & adj_pvalue < (0.0001))  %>%
  group_by(comparison) %>% summarise(n = n()) 


TEclass_counts <- kruskal_results.TE %>%
  group_by(TEclass, significant) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = significant, values_from = count, values_fill = 0) %>%
  rename(No = "No", Yes = "Yes")




TEclass_stats <- TEclass_counts %>%
  rowwise() %>%
  mutate(
    p_value = fisher.test(matrix(c(No, Yes, 839 , 5968), nrow = 2))$p.value
  ) %>%
  ungroup() %>%
  mutate(ratio = Yes/No) %>%
  arrange(-ratio) %>%
  mutate(average =839/5968 ) %>%
  mutate(median = median(ratio))


methylationData.3.QC.TE %>%
  distinct(TE_ID2)

methylationData.3.QC.TE %>% filter(TE_ID2 == "8097") %>%
  ggplot(aes(x = start, y = percent, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.1) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(75,100)) +
  labs(
    title = "Methylation across TE8097 (Plavaka)",
    x = "Position in TE8097",
    y = "Percent methylated"
  ) 

kruskal_results.TE.significant %>% slice_head(n = 5)  -> 
  top5.te
methylationData.3.QC.TE %>% 
  filter(TE_ID2 %in%  top5.te$TE_ID2) %>%
  ggplot(aes(x = start, y = percent, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.4) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(85,100)) +
  labs(
    title = "Methylation across different TEs (top 5 significant)",
    x = "Position in TE8097",
    y = "Percent methylated"
  ) + facet_wrap("TE_ID2",nrow = 1, scales = "free_x")   +
  theme_light() -> examples.methylation
  

methylationData.3.QC.TE %>% 
  filter(TE_ID2 %in%  top5.te$TE_ID2) %>%
  ggplot(aes(x = start, y = depth, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.4) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(0,40))  +
  labs(
    title = "Depth across different TEs (top 5 significant)",
    x = "Position in TE",
    y = "Depth"
  ) + facet_wrap(c("TE_ID2","TEclass"),nrow = 1, scales = "free_x")   +
  theme_light()  -> examples.depth


methylationData.3.QC.TE %>% 
  filter(TE_ID2 == "8097") %>%
  ggplot(aes(x = start, y = percent, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.4) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(80,100))  +
  labs(
    title = "Depth across different Samples",
    x = "Position in TE",
    y = "Depth"
  ) + facet_wrap(c("Sample"),nrow = 1, scales = "free_x")   +
  theme_light() 

methylationData.3.QC.TE %>% 
  filter(TE_ID2 == "8097") %>%
  ggplot(aes(x = start, y = percent, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.4) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(80,100))  +
  labs(
    title = "Depth across different Samples",
    x = "Position in TE",
    y = "Depth"
  )  +
  theme_light() 









```


#### Linear mixed model
```{r}
methylationData.3.QC.TE %>%
  group_by(TE_ID2) %>%
  summarise(
    p_value = tryCatch({
      kruskal.test(percent ~ Tissue, data = cur_data())$p.value
    }, error = function(e) NA),n = n()
  ) %>%
  mutate(comparison = "Tissue") %>%
  mutate(adj_pvalue = p.adjust(p_value, method = "BH")) %>%
  inner_join(TE.info) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  kruskal_results.TE

methylationData.3.QC.TE %>% distinct(TE_ID2) ->
  TE.unique

methylationData.3.QC.TE %>%
  select(TE_ID2,Sample,Tissue,start,percent) %>%
  ungroup() %>%
  group_by(TE_ID2,Tissue,start) %>%
    summarise(n = n(), methylation = mean(percent)) ->
  TE.summary.all
  
TE.summary.all %>%
  group_by(TE_ID2,Tissue) %>%
  summarise(n = n()) %>%
  ungroup %>%
  filter(n >=10) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n >=2) ->
  TE.filtered

TE.summary.all %>% filter(TE_ID2 %in% TE.filtered$TE_ID2) %>%
  ungroup() %>%
  distinct(TE_ID2,start) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n > 10) ->
  TE.filtered.nCpG.nObs
  



```



```{r}

library(lme4)
#install.packages("lmerTest")
library(lmerTest)   # adds p-values


dist.list.Tissue = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Tissue,start,percent) ->
      mat
     model <- lmer(percent ~ Tissue + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Tissue[[TE]] = TEinfo
       
}

dist.list.Location = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Location,start,percent) ->
      mat
     model <- lmer(percent ~ Location + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Location[[TE]] = TEinfo
       
}

dist.list.Rep = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Replicate2,start,percent) ->
      mat
     model <- lmer(percent ~ Replicate2 + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Rep[[TE]] = TEinfo
       
}


TEspvalues.Tissue = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Tissue[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Tissue[[TE]][["Name"]],
  Test = "Tissue"
  ) ->
  TEspvalues.Tissue[[as.character(TE)]]
}

TEspvalues.Location = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Location[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Location[[TE]][["Name"]],
  Test = "Location"
  ) ->
  TEspvalues.Location[[as.character(TE)]]
}

TEspvalues.Rep = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Rep[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Rep[[TE]][["Name"]],
  Test = "Rep"
  ) ->
  TEspvalues.Rep[[as.character(TE)]]
}



bind_rows(TEspvalues.Tissue) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Tissue


bind_rows(TEspvalues.Location) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Location

bind_rows(TEspvalues.Rep) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Rep

lmm_results.TE = bind_rows(lmm_results.TE.Tissue,lmm_results.TE.Location,lmm_results.TE.Rep )


lmm_results.TE %>% ggplot(aes(x = pvalue, color = Test)) +
  geom_density( adjust = 1/10) +
  labs(
    title = "Density plot",
    x = "p-value",
    y = "Density",
    shape = "Class"
  ) + theme_light() ->
  TE.pvalue.density.plot
ggsave("~/MarasmiusMethylationRNA/figures/Figure3/3A.kruskal.distribution.plot.pdf",plot = TE.pvalue.density.plot,width = 180, height = 180, units = "mm", dpi = 300)


lmm_results.TE %>% arrange(adj_pvalue) %>%
  filter(significant == "Yes") ->
  lmm_results.TE.significant
lmm_results.TE.significant %>% 
  group_by(Test) %>%
  summarise(numberOfSignificant = n()) ->
  lmm_results.TE.significant.summary

```

###  Aalysis of  significant TEs  
```{r }    
lmm_results.TE.Tissue %>% filter(significant == "Yes") %>%
  arrange(adj_pvalue) ->
  lmm_results.TE.Tissue.significant

TE_tissue_DE = intersect(lmm_results.TE.Tissue.significant$TE_ID2,as.numeric( rownames(methylationData.3.QC.TE.wide))) 

methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]->

methylationData.3.QC.TE.sig.wide 





```
#### Cluster of TEs
##### Hierarchical clustering with correct number of clusters
```{r}


## Cluster analysis
dist_matrix <- dist(methylationData.3.QC.TE.sig.wide, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
library(NbClust)
nb <- NbClust(methylationData.3.QC.TE.sig.wide, distance="euclidean", min.nc=2, max.nc=10, method="ward.D2")


wss <- numeric(10)
for (k in 2:10) {
  clusters <- cutree(hc, k)
  wss[k] <- sum(sapply(unique(clusters), function(c) {
    pts <- methylationData.3.QC.TE.sig.wide[clusters == c, ]
    sum(scale(pts, scale=FALSE)^2)
  }))
}

plot(1:10, wss, type="b", xlab="k", ylab="Within-cluster SS")




nrOfClusters = nb$Best.nc[[1]]

nrOfClusters = 3
clusters <- cutree(hc, nrOfClusters)


clusters %>%  as.data.frame() %>%
  rownames_to_column(var = "TE_ID2") %>%
  rename(cluster = ".") ->
  clusters




```


#####  Visualisation of 3 clusters

```{r}


clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(cluster, Tissue) %>%
  summarise(mean = mean(percent), sd = sd(percent))->
  clusters.Tissue.summary


clusters.Tissue.summary %>% group_by(cluster) %>%
  summarise(overallMean = mean(mean)) %>%
  arrange(overallMean) %>%
  mutate(clusterOrder = as.factor(row_number()))->
  cluster.order
  



clusters %>%  
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  ungroup() %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n()) %>%
  arrange(cluster, -nrOfTEs)%>%
  group_by(cluster) %>%
  slice_head(n = 5)->
  clusters.count.summary


clusters.Tissue.summary %>%
  mutate(clusterName = "Low Methylation") %>%
  mutate(clusterName = replace(clusterName, 
                               grepl(1, cluster),
                               "Medium Methylation")) %>% 
  
  mutate(clusterName = replace(clusterName, 
                               grepl(3, cluster),
                               "High Methylation")) %>%
  mutate(clusterName = factor(clusterName, 
                              levels = c("Low Methylation",
                                         "Medium Methylation",
                                         "High Methylation"))) %>%
  inner_join(clusters.count.summary)->
clusters.summary

clusters.Tissue.summary %>%
  inner_join(clusters.count.summary) %>%
  inner_join(cluster.order)->
clusters.summary
  


clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(TE_ID2,Tissue,Sample,cluster ) %>%
  summarise(percent = mean(percent) )%>%
  inner_join(clusters.summary) %>%
  ggplot(aes(x = clusterOrder, y = percent, fill = Tissue)) +
  geom_boxplot() +
  theme_light() +
  labs(
    title = "Methylation levels per TE cluster",
    x = "TE Cluster",
    y = "Percent Methylation"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )+
  scale_fill_manual(values = colors) -> distributionPerClusterPlot
   
     

ha = HeatmapAnnotation(
    cluster = as.factor(clusters$cluster)
)

distributionPerClusterPlot

pdf("~/MarasmiusMethylationRNA/figures/Figure3/TE.methylation.clustered.heatmap.pdf",
    width = 8, height = 6)
ComplexHeatmap::Heatmap(t(methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]),
                        show_column_names = FALSE,
                        show_column_dend = TRUE, 
                        cluster_columns = hc,
                        column_split = nrOfClusters,
                        top_annotation = ha,
                        cluster_rows =TRUE,  split = 2
) 

dev.off()
distributionPerClusterPlot


clusters %>%   
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n())  %>%
  inner_join(cluster.order) %>%
  filter(TEclass %in% clusters.count.summary$TEclass) %>%
  ggplot(aes(x = clusterOrder, y = nrOfTEs, fill = TEclass)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_light() +
  labs(
    title = "TE class distribution per TE cluster",
    x = "TE Cluster",
    y = "Number of TEs"
  ) -> distribution


cluster.order

clusters%>%   
  inner_join(clusters.count.summary) %>%   
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n())  %>%
  inner_join(cluster.order) %>%
  ggplot(aes(x = clusterOrder, y = nrOfTEs)) +
  geom_bar(stat = "identity") +
  theme_light() +
  labs(
    title = "TE class distribution per TE cluster",
    x = "TE Cluster",
    y = "Number of TEs"
  ) -> distribution2



distributionPerClusterPlot/distribution2/distribution




```





###  mRNA analysis

```{r }
TEdf %>% 
  mutate(motifStart = as.numeric(motifStart)) %>%
  mutate(motifStop = as.numeric(motifStop)) %>%
  separate(TEclass, into=c("TEclass","nr"), sep = "-") ->
  TEdf
TEdf %>%
  group_by(TEclass) %>%
  summarise(motifStart = min(motifStart), motifStop = max(motifStop)) %>%
  arrange(-motifStop)  %>%
  filter(motifStop>500) %>%
  select(TEclass) %>%
  inner_join(TEdf)->
  TEdf.long

TEdf.long %>% select(TEclass,seqid,TE_ID2) -> 
  TE.info

TE.info 
  

annotation.mRNA.df %>% select(locus_tag,product)%>% 
  distinct() %>% 
  group_by(locus_tag) %>%
  slice_head(n = 1) ->
  annotation.mRNA.df.short 


methylationData.3.QC %>% 
  filter(location == "GeneBody")  %>%
  inner_join(annotation.mRNA.df.short) ->
  methylationData.3.QC.GB


methylationData.3.QC.GB %>%
  mutate(GB_ID = paste(locus_tag,Location,Replicate2,Tissue,sep = "_")) ->
  methylationData.3.QC.GB

GB_ID = methylationData.3.QC.GB %>% distinct(GB_ID, locus_tag,Location,Replicate2,Tissue)





```


#### Distance to nearest TE analysis
```{r }


methylationData.3.QC.GB %>%
  group_by(GB_ID) %>%
  summarise( percent = mean(percent), nrOfExamples = n()) %>%
  inner_join(GB_ID) %>%
  left_join(GeneBody_TE_distance)->
  methylationData.3.QC.GB.summary


methylationData.3.QC.GB.summary %>%
  group_by(GB_ID) %>%
  slice_head(n = 1) %>%
  ungroup() ->
  methylationData.3.QC.GB.summary

methylationData.3.QC.GB.summary %>%
  mutate(
    location = case_when(
      distance < -500                ~ "Far Upstream",
      distance < 0                   ~ "Near Upstream",
      distance == 0                  ~ "Within",
      distance > 0 & distance < 500  ~ "Close",
      distance >= 500 & distance < 2000  ~ "Proximal",
      distance >= 2000 & distance < 5000 ~ "Intermediate",
      TRUE                           ~ "Long-range"
    ),
    location = factor(location, levels = c(
      "Far Upstream", "Near Upstream", "Within", 
      "Close", "Proximal", "Intermediate", "Long-range"
    ))
  ) %>%
  mutate(
    methylationLevel = case_when(
      percent < 5                ~ "None",
      percent >= 5 & percent < 25  ~ "Low",
      percent >= 25 & percent < 75  ~ "Intermediate",
      percent >= 75 & percent < 95 ~ "High",
      TRUE                           ~ "All"
    ),
    methylationLevel = factor(methylationLevel, levels =rev( c(
      "None", "Low", "Intermediate", "High","All"
    )))
  ) ->
  methylationData.3.QC.GB.summary





methylationData.3.QC.GB.summary%>%
  inner_join(TE.info) ->
  methylationData.3.QC.GB.summary
  

  


methylationData.3.QC.TE %>%
  select(TE_ID2,Location,Replicate2,Tissue, percent) %>%
  rename(TE_percent = percent) %>%
  group_by(TE_ID2,Location,Replicate2,Tissue) %>%
  summarise(TE_percent = mean(TE_percent))->
  methylationData.3.QC.TE.comparison




methylationData.3.QC.GB.summary %>%
  inner_join(methylationData.3.QC.TE.comparison, by = c("TE_ID2","Location","Replicate2","Tissue"))  %>%
#  filter(TE_percent >1 & percent >1) %>%
# filter(distance < 500) %>%
  ggplot(aes(x = TE_percent, y = percent))  + 
  geom_hex(bins = 30) +  # Adjust 'bins' to change resolution
scale_fill_viridis_c(trans = "log10") +
  theme_minimal() +
  labs(title = "Point Density Map", 
       fill = "Count of Points")  +
  facet_wrap("location") ->
  TE_GeneBody_methylation_distance_less500.location.plot
  
library(here)
ggsave(here("figures","Figure4","TE_GeneBody_methylation_distance_less500.pdf"),
       plot = TE_GeneBody_methylation_distance_less500.location.plot,
       width = 8, height = 6, units = "in", dpi = 300)


methylationData.3.QC.GB.summary %>%
  inner_join(methylationData.3.QC.TE.comparison, by = c("TE_ID2","Location","Replicate2","Tissue"))  %>%
#  filter(TE_percent >1 & percent >1) %>%
# filter(distance < 500) %>%
  ggplot(aes(x = TE_percent, y = percent))  + 
  geom_hex(bins = 30) +  # Adjust 'bins' to change resolution
scale_fill_viridis_c(trans = "log10") +
  theme_minimal() +
  labs(title = "Point Density Map", 
       fill = "Count of Points")  +
  facet_grid(location~TEclass) ->
  TE_GeneBody_methylation.location.TEclass.plot


ggsave(here("figures","Figure4","TE_GeneBody_methylation.location.TEclass.plot.pdf"),
       plot = TE_GeneBody_methylation.location.TEclass.plot,
       width = 20, height = 12, units = "in", dpi = 300)







methylationData.3.QC.GB.summary%>%
  ggplot(aes(x = as.factor(location), fill = methylationLevel)) +
  geom_bar(alpha = 0.7, position = "fill") +
  theme_minimal() +
  labs(title = "Point Density Map", 
       fill = "Count of Points")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "RdYlBu") -> distribution.methylationLevel

methylationData.3.QC.GB.summary%>%
  ggplot(aes(x = as.factor(location), fill = methylationLevel)) +
  geom_bar(alpha = 0.7, position = "fill") +
  theme_minimal() +
  labs(title = "Point Density Map", 
       fill = "Count of Points")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "RdYlBu") -> distribution.methylationLevel





all <- methylationData.3.QC.GB.summary %>% mutate(methylationLevel = "All")

bind_rows(methylationData.3.QC.GB.summary,all)%>%
  ggplot(aes(x = log10(distance+1), color = methylationLevel)) +
  geom_density(alpha = 0.7) +
  theme_minimal() 



methylationData.3.QC.GB.summary %>% 
  group_by(location,methylationLevel) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = methylationLevel, values_from = n, values_fill = 0)->
  table.methylationLevel.location






```


####PCA analysis

```{r }


methylationData.3.QC.GB.summary %>%
  group_by(TE_ID2,Tissue) %>%
  summarise(nrOfExamples = n()) -> methylationData.3.QC.GB.summary.2

methylationData.3.QC.GB.summary.2 %>%
  group_by(Tissue) %>%
  summarise(n = max(nrOfExamples)) %>%
  inner_join(methylationData.3.QC.GB.summary.2) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n == 3)  %>%
  select(TE_ID2)->
  GB.covered


GB.covered
  

methylationData.3.QC.GB.summary %>%
  inner_join(TE.covered) %>%
  select(locus_tag,TE_ID2,Location,Replicate2,Tissue, percent) %>%
  pivot_wider(values_from = "percent",names_from = c("Location","Replicate2","Tissue")) %>%
  column_to_rownames(var = "locus_tag")%>%
  drop_na()->                     # Removes any row with at least one NA->
  methylationData.3.QC.GB.wide
  
  
  methylationData.3.QC.GB.summary %>%ungroup() %>%
    group_by(locus_tag,Location,Replicate2,Tissue) %>%
    summarise(n = n()) %>%
    arrange(-n)

## PCA analysis
PCAresult = PCAtools::pca(methylationData.3.QC.GB.wide[,-1],center = TRUE)
rot = PCAresult$rotated[,1:6]

## Cluster analysis
dist_matrix <- dist(rot, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")

#library(NbClust)
#nb <- NbClust(rot, distance="euclidean", min.nc=2, max.nc=6, method="ward.D2")
#nb$Best.nc


#wss <- numeric(10)
#for (k in 2:10) {
#  clusters <- cutree(hc, k)
#  wss[k] <- sum(sapply(unique(clusters), function(c) {
#    pts <- rot[clusters == c, ]
#    sum(scale(pts, scale=FALSE)^2)
#  }))
#}

#plot(1:10, wss, type="b", xlab="k", ylab="Within-cluster SS")


## Variance
PCA.stats = data.frame(variance = PCAresult$variance, 
                       PC = factor(names(PCAresult$variance), levels = names(PCAresult$variance)))
ggplot(PCA.stats, aes(x = PC, y = variance) )+
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(
    title = "PC variance",
    x = "PC",
    y = "Variance"
  )+theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  ) -> variation


ggsave("~/MarasmiusMethylationRNA/figures/figure3/GB.PCA.variance.plot.pdf",
       plot = variation, width = 10, height = 8, units = "in")


PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary) %>%
  ggplot(aes(x = PC1, y = PC2, color =Tissue, shape = Location )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  scale_color_manual(values = colors) +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  )-> PCA.plot

PCA.plot +variation
  
PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary)  %>%
  ggplot(aes(x = PC1, y = PC2, color =Tissue, shape = Location,label =sampleID )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  scale_color_manual(values = colors) +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  ) + geom_text_repel() -> PCA.plot.labels



PCA.plot.labels


breaks1 <- seq(from = -100, to = 300, length.out = 4)
library(RColorBrewer)
colors1 = colorRampPalette(rev(brewer.pal(n = 7, name =
  "RdYlBu")))(4)
myColor2 <-circlize::colorRamp2(breaks = breaks1,colors =colors1)


library(ComplexHeatmap)

pdf("~/MarasmiusMethylationRNA/figures/Figure3/TE.methylation.PCA.heatmap.pdf",
    width = 1800, height = 1200)
ComplexHeatmap::Heatmap(
  PCAresult$rotated[1:3] ,
  col = myColor2,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(
      x, y, width = width, height = height,
      gp = gpar(col = "darkgrey", fill = NA, lwd = 1) # Add gray borders
    )
  }
) 
dev.off()



# Determine the optimal number of clusters
# Plot the dendrogram
plot(hc, main = "Hierarchical Clustering Dendrogram", sub = "", xlab = "", ylab = "Height", hang = -1, cex = 0.6)

# Cut the dendrogram into 3 clusters
clusters <- cutree(hc, k = 4)

# Add cluster assignments to the data frame
rot$cluster <- as.factor(clusters)





 rot = rot %>% rownames_to_column("IDsample") 
 methylationData.3.QC.annotation.distribution %>% 
    select(Tissue,Location,Replicate2,methylation,location) %>%
      mutate(IDsample = paste(Location,Replicate2,Tissue,sep = "_")) %>%
    inner_join(rot) %>%
    pivot_wider(values_from = "methylation", names_from = "location" )  %>%
    ggplot(aes(x = GeneBody, y = TE, shape = Tissue,color = cluster)) + geom_point()+
    theme_light()+
    geom_text_repel(aes(label = IDsample), color = "black") -> fig3B.clustered
     

 PCAresult$rotated  %>%
  rownames_to_column(var = "sampleID") %>%
  inner_join(sampleInfo.TE.summary)  %>%
  inner_join(rot) %>%
  ggplot(aes(x = PC1, y = PC2, color =cluster, shape = Tissue,label =sampleID )) +
  geom_point(size = 3, alpha = 0.7) + # Adjust point size
  scale_shape_manual(values = c(16, 17, 15, 18)) + # Custom shapes
  theme_minimal() +
  labs(
    title = "Scatter Plot with Four Classes",
    x = "PC1 (29 % variance)",
    y = "PC2 (8.6 % variance )",
    shape = "Class"
  ) +geom_text_repel(aes(label = sampleID), color = "black")  -> 
  
  PCA.plot.clustered
 PCA.plot.clustered
 
 
 GO annotation on the first loading 
```


#### Linear mixed model


```{r}
TE.summary.all %>%
  group_by(TE_ID2,Tissue) %>%
  summarise(n = n()) %>%
  ungroup %>%
  filter(n >=10) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n >=2) ->
  TE.filtered

TE.summary.all %>% filter(TE_ID2 %in% TE.filtered$TE_ID2) %>%
  ungroup() %>%
  distinct(TE_ID2,start) %>%
  group_by(TE_ID2) %>%
  summarise(n = n()) %>%
  filter(n > 10) ->
  TE.filtered.nCpG.nObs
  

```



```{r}

library(lme4)
#install.packages("lmerTest")
library(lmerTest)   # adds p-values





dist.list.Tissue = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Tissue,start,percent) ->
      mat
     model <- lmer(percent ~ Tissue + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Tissue[[TE]] = TEinfo
       
}

dist.list.Location = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Location,start,percent) ->
      mat
     model <- lmer(percent ~ Location + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Location[[TE]] = TEinfo
       
}

dist.list.Rep = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  TEinfo = list()
    methylationData.3.QC.TE %>%
      filter(TE_ID2 == TE) %>%
      select(Sample,Replicate2,start,percent) ->
      mat
     model <- lmer(percent ~ Replicate2 + (1 | start), data = mat)
       pval  <- anova(model)$`Pr(>F)`[1]
     TEinfo[["Name"]] = TE
     TEinfo[["data"]] = mat
     TEinfo[["mixedLinearModel"]] = model
     TEinfo[["pvalue"]] = pval
     
     dist.list.Rep[[TE]] = TEinfo
       
}


TEspvalues.Tissue = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Tissue[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Tissue[[TE]][["Name"]],
  Test = "Tissue"
  ) ->
  TEspvalues.Tissue[[as.character(TE)]]
}

TEspvalues.Location = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Location[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Location[[TE]][["Name"]],
  Test = "Location"
  ) ->
  TEspvalues.Location[[as.character(TE)]]
}

TEspvalues.Rep = list()
for(TE in TE.filtered.nCpG.nObs$TE_ID2){
  data.frame(
  pvalue = dist.list.Rep[[TE]][["pvalue"]],
  TE_ID2 = dist.list.Rep[[TE]][["Name"]],
  Test = "Rep"
  ) ->
  TEspvalues.Rep[[as.character(TE)]]
}



bind_rows(TEspvalues.Tissue) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Tissue


bind_rows(TEspvalues.Location) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Location

bind_rows(TEspvalues.Rep) %>%
  mutate(adj_pvalue = p.adjust(pvalue, method = "BH")) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, adj_pvalue < (0.0001), "Yes")) ->
  lmm_results.TE.Rep

lmm_results.TE = bind_rows(lmm_results.TE.Tissue,lmm_results.TE.Location,lmm_results.TE.Rep )


lmm_results.TE %>% ggplot(aes(x = pvalue, color = Test)) +
  geom_density( adjust = 1/10) +
  labs(
    title = "Density plot",
    x = "p-value",
    y = "Density",
    shape = "Class"
  ) + theme_light() ->
  TE.pvalue.density.plot
ggsave("~/MarasmiusMethylationRNA/figures/Figure3/3A.kruskal.distribution.plot.pdf",plot = TE.pvalue.density.plot,width = 180, height = 180, units = "mm", dpi = 300)


lmm_results.TE %>% arrange(adj_pvalue) %>%
  filter(significant == "Yes") ->
  lmm_results.TE.significant
lmm_results.TE.significant %>% 
  group_by(Test) %>%
  summarise(numberOfSignificant = n()) ->
  lmm_results.TE.significant.summary

```

###  Aalysis of  significant TEs  
```{r }    
lmm_results.TE.Tissue %>% filter(significant == "Yes") %>%
  arrange(adj_pvalue) ->
  lmm_results.TE.Tissue.significant

TE_tissue_DE = intersect(lmm_results.TE.Tissue.significant$TE_ID2,as.numeric( rownames(methylationData.3.QC.TE.wide))) 

methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]->

methylationData.3.QC.TE.sig.wide 





```
#### Cluster of TEs
##### Hierarchical clustering with correct number of clusters
```{r}


## Cluster analysis
dist_matrix <- dist(methylationData.3.QC.TE.sig.wide, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
library(NbClust)
nb <- NbClust(methylationData.3.QC.TE.sig.wide, distance="euclidean", min.nc=2, max.nc=10, method="ward.D2")


wss <- numeric(10)
for (k in 2:10) {
  clusters <- cutree(hc, k)
  wss[k] <- sum(sapply(unique(clusters), function(c) {
    pts <- methylationData.3.QC.TE.sig.wide[clusters == c, ]
    sum(scale(pts, scale=FALSE)^2)
  }))
}

plot(1:10, wss, type="b", xlab="k", ylab="Within-cluster SS")




nrOfClusters = nb$Best.nc[[1]]

nrOfClusters = 3
clusters <- cutree(hc, nrOfClusters)


clusters %>%  as.data.frame() %>%
  rownames_to_column(var = "TE_ID2") %>%
  rename(cluster = ".") ->
  clusters




```


#####  Visualisation of 3 clusters

```{r}


clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(cluster, Tissue) %>%
  summarise(mean = mean(percent), sd = sd(percent))->
  clusters.Tissue.summary


clusters.Tissue.summary %>% group_by(cluster) %>%
  summarise(overallMean = mean(mean)) %>%
  arrange(overallMean) %>%
  mutate(clusterOrder = as.factor(row_number()))->
  cluster.order
  



clusters %>%  
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  ungroup() %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n()) %>%
  arrange(cluster, -nrOfTEs)%>%
  group_by(cluster) %>%
  slice_head(n = 5)->
  clusters.count.summary


clusters.Tissue.summary %>%
  mutate(clusterName = "Low Methylation") %>%
  mutate(clusterName = replace(clusterName, 
                               grepl(1, cluster),
                               "Medium Methylation")) %>% 
  
  mutate(clusterName = replace(clusterName, 
                               grepl(3, cluster),
                               "High Methylation")) %>%
  mutate(clusterName = factor(clusterName, 
                              levels = c("Low Methylation",
                                         "Medium Methylation",
                                         "High Methylation"))) %>%
  inner_join(clusters.count.summary)->
clusters.summary

clusters.Tissue.summary %>%
  inner_join(clusters.count.summary) %>%
  inner_join(cluster.order)->
clusters.summary
  


clusters %>%
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  group_by(TE_ID2,Tissue,Sample,cluster ) %>%
  summarise(percent = mean(percent) )%>%
  inner_join(clusters.summary) %>%
  ggplot(aes(x = clusterOrder, y = percent, fill = Tissue)) +
  geom_boxplot() +
  theme_light() +
  labs(
    title = "Methylation levels per TE cluster",
    x = "TE Cluster",
    y = "Percent Methylation"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )+
  scale_fill_manual(values = colors) -> distributionPerClusterPlot
   
     

ha = HeatmapAnnotation(
    cluster = as.factor(clusters$cluster)
)

distributionPerClusterPlot
ComplexHeatmap::Heatmap(t(methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]),
                        show_column_names = FALSE,
                        show_column_dend = TRUE, 
                        cluster_columns = hc,
                        column_split = nrOfClusters,
                        top_annotation = ha,
                        cluster_rows =TRUE,  split = 2
) 

distributionPerClusterPlot


clusters %>%   
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n())  %>%
  inner_join(cluster.order) %>%
  filter(TEclass %in% clusters.count.summary$TEclass) %>%
  ggplot(aes(x = clusterOrder, y = nrOfTEs, fill = TEclass)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_light() +
  labs(
    title = "TE class distribution per TE cluster",
    x = "TE Cluster",
    y = "Number of TEs"
  ) -> distribution


cluster.order

clusters%>%   
  inner_join(clusters.count.summary) %>%   
  mutate(TE_ID2 = as.double(TE_ID2)) %>%
  inner_join(methylationData.3.QC.TE) %>%
  distinct(TE_ID2,cluster,TEclass) %>%
  group_by(cluster,TEclass) %>%
  summarise(nrOfTEs = n())  %>%
  inner_join(cluster.order) %>%
  ggplot(aes(x = clusterOrder, y = nrOfTEs)) +
  geom_bar(stat = "identity") +
  theme_light() +
  labs(
    title = "TE class distribution per TE cluster",
    x = "TE Cluster",
    y = "Number of TEs"
  ) -> distribution2



distributionPerClusterPlot/distribution/distribution2




```





```{r}



ComplexHeatmap::Heatmap(t(methylationData.3.QC.TE.wide[as.character(TE_tissue_DE),]),
                        show_column_names = FALSE,
                        show_column_dend = FALSE, column_split = 5,
                        cluster_rows =  hc,split = 4
) -> HeatMap.TE.significant




variation+ PCA.plot + fig3B.clustered + HeatMap.TE.significant +
  plot_layout(ncol = 2, nrow = 2, guides = "collect") +
  plot_annotation(tag_levels = 'A')-> figure3


ggsave(PCA.plot.clustered)("~/MarasmiusMethylationRNA/figures/Figure3/Figure3.TE.methylation.PCA.2clusters.pdf",
       plot = figure3, width = 12, height = 10, units = "in", dpi = 300,
       plot = fig3B.clustered)

```



```{r }
lmm_results.TE.Tissue.significant %>% 
  slice_head(n = 20) %>%inner_join(methylationData.3.QC.TE) -> examples




examples %>% 
  mutate(TEclass_ID = paste(TEclass, TE_ID2, sep = "\n")) %>%
  arrange(TE_ID2,TEclass) %>%
  filter(TEclass != "Gypsy") %>%
  ggplot(aes(x = Tissue, y = percent)) + geom_violin() + facet_wrap(facets = "TEclass_ID") 


TEclass_stats %>%
  mutate(TEclass = factor(TEclass,levels = TEclass_stats$TEclass)) %>%
  ggplot(aes(x= TEclass, y = ratio)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Bar plot
  theme_minimal() +  # Minimal theme
  labs(x = "TE type", y = "Ratio (Sig/nonSig)") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )

TEclass_stats %>%
  mutate(TEclass = factor(TEclass,levels = TEclass_stats$TEclass)) %>%
  mutate(Total = No+Yes) %>%
  ggplot(aes(x= TEclass, y = Total)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Bar plot
  theme_minimal() +  # Minimal theme
  labs(x = "TE type", y = " Number of TEs") +
  theme(
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank() ,   # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  ) -> countPlot

TEclass_stats %>%
  mutate(TEclass = factor(TEclass,levels = TEclass_stats$TEclass)) %>%
  pivot_longer(names_to = "Significant", values_to = "Count", cols = c(No, Yes))  %>%
  mutate(Significant = factor(Significant, levels = c("Yes","No"))) %>%
  group_by(TEclass) %>%
  mutate(Fraction = Count / sum(Count)) %>% 
  ggplot(aes(x= TEclass, y = Fraction, fill = Significant)) +
  geom_bar(stat = "identity") +  # Bar plot
  theme_minimal() +  # Minimal theme
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  ) -> ratioPlot


library(patchwork)
countPlot/ratioPlot



```



#### Get where the methylation is the highest
```{r}

methylationData.3.QC.TE %>%
  ungroup() %>% 
  group_by(TE_ID2,Tissue) %>%
  summarise(mean = mean(percent)) ->
  methylationData.3.QC.TE.summary

methylationData.3.QC.TE.summary %>%
  group_by(TE_ID2) %>%                        # Group by TE_ID2
  mutate(highest_tissue = Tissue[which.max(mean)]) %>%  # Find the tissue with the max mean
  ungroup() %>%
  inner_join(kruskal_results.TE.significant) %>%
  pivot_wider(values_from = "mean", names_from = "Tissue") %>%
  arrange(p_value) %>% inner_join(TE.info) ->
  TE.class.info


methylationData.3.QC.TE.summary %>%
  group_by(TE_ID2) %>%                        # Group by TE_ID2
  mutate(highest_tissue = Tissue[which.max(mean)]) %>%  # Find the tissue with the max mean
  ungroup()  %>%
  distinct(TE_ID2,highest_tissue)  %>%
  inner_join(kruskal_results.TE.significant) %>%
  inner_join(TE.info) %>%
  dplyr::count(highest_tissue,TEclass) %>%
  pivot_wider(names_from = "highest_tissue",
              values_from = "n",
              values_fill = 0) %>%
  arrange(-lamella) %>%
  inner_join(TEclass_stats)  -> TEclass_stats.2




TEclass_stats.2 %>%
  mutate(TEclass = factor(TEclass,levels = TEclass_stats$TEclass)) %>%
  pivot_longer(names_to = "Significant", values_to = "Count", cols = c(lamella,spore,stipe,No))  %>%
  mutate(Significant = factor(Significant, levels = c("spore","lamella","stipe","No"))) %>%
  group_by(TEclass) %>%
  mutate(Fraction = Count / sum(Count)) %>% 
  ggplot(aes(x= TEclass, y = Fraction, fill = Significant)) +
  geom_bar(stat = "identity") +  # Bar plot
  theme_minimal() +  # Minimal theme
  labs(x = "TE type", y = " Fraction") +
  scale_fill_manual(values = colors) + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )-> ratioPlot.2
  
countPlot/ratioPlot.2



```

#### Check coverage correlation to significant methylation pattern (Defence approach)
```{r}
TE.class.info %>% select(TE_ID2,highest_tissue) -> TE.sig.tissue

kruskal_results.TE %>% inner_join(TEdf) %>%
  group_by(seqid,significant) %>%
  summarise(n = n()) -> TE.position.distribution
  
kruskal_results.TE %>%
  inner_join(TEdf.long) %>%
  ungroup() %>%
  group_by(seqid) %>%
  summarise(n = n()) %>%
  filter(n >100) %>%
  ungroup() %>%
  select(seqid) %>%
  inner_join(kruskal_results.TE) %>%
  inner_join(TEdf.long)  %>%
  left_join(TE.sig.tissue) %>%
  mutate(highest_tissue = replace(highest_tissue,is.na(highest_tissue),"No")) %>%
  mutate(highest_tissue = factor(highest_tissue,
                                 levels = c("spore","lamella","stipe","No")))  ->
  kruskal_results.TE.filtered


kruskal_results.TE.filtered %>%
  ggplot(aes(x = seqid, y = start, color = highest_tissue )) + 
    geom_point(size = 0.6, alpha= 0.6) + facet_wrap(facets = "TEclass")+
    theme_minimal() +  # Minimal theme
    scale_color_manual(values = colors)  + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5, size = 8),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )

```

##### Gypsy
```{r}
     
kruskal_results.TE.filtered %>% filter(TEclass =="Gypsy") ->
  kruskal_results.TE.filtered.Gypsy



kruskal_results.TE.filtered.Gypsy %>%
  ggplot(aes(x =highest_tissue , y = start, color = highest_tissue )) + 
    geom_point(size = 0.6, alpha= 0.4) +
  facet_wrap(facets = "seqid")+  
  theme_minimal() +  # Minimal theme
  scale_color_manual(values = colors)  + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )
    



kruskal_results.TE.filtered.Gypsy %>%
   group_by(seqid,highest_tissue) %>%
   summarise(n = n())  %>% 
  ggplot(aes(x= seqid, y = n, fill = highest_tissue)) +
  geom_bar(stat = "identity") +  # Bar plot
  theme_minimal() +  # Minimal theme
  labs(x = "Chromosome", y = " Number of TEs") +
  scale_fill_manual(values = colors) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )
  
 
 kruskal_results.TE.filtered.Gypsy %>%
   ggplot(aes(x = start, color =highest_tissue) )+ facet_wrap(facets = "seqid")+  # Bar plot
  theme_minimal() +
    scale_color_manual(values = colors)+ geom_density( adjust = 0.2) +
   geom_density(color = "black", adjust = 0.2)+ scale_y_continuous(
    limits = c(0, 0.000003), # Set the y-axis range to show only 4 to 8
    expand = c(0, 0)  # Remove padding at the edges
  ) 
 
 kruskal_results.TE.filtered.Gypsy %>%
   ggplot(aes(y = start,x = seqid, color =highest_tissue) )+ 
   theme_minimal() +
   scale_color_manual(values = colors)+ 
   geom_violin(position="identity",adjust = 0.2, color = "black")+
   geom_violin(position="identity",adjust = 0.2,alpha=0.0)+ 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate and adjust x-axis labels
    axis.title.x = element_text(size = 12, face = "bold"),  # Format x-axis title
    axis.title.y = element_text(size = 12, face = "bold")   # Format y-axis title
  )-> violinPlot
   
   
   scale_y_continuous(
    limits = c(0, 0.000003), # Set the y-axis range to show only 4 to 8
    expand = c(0, 0)  # Remove padding at the edges
  ) 

 
 
 

```

###  mRNA analysis

####  Non-Parametric Test (Kruskal-Wallis)
```{r }
  
methylationData.3.QC %>%
  filter(location == "GeneBody")  ->
  methylationData.3.QC.mRNA




methylationData.3.QC.mRNA %>%
  group_by(Name) %>%
  summarise(
    p_value = tryCatch({
      kruskal.test(percent ~ Tissue, data = cur_data())$p.value
    }, error = function(e) NA)
  ) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, p_value < 0.0000001, "Yes")) ->
  kruskal_results.mRNA



# Filter significant results
kruskal_results.mRNA %>%
  filter(!is.na(p_value) & p_value < 0.00001)%>%
  arrange(p_value) ->
  kruskal_results.mRNA.sig

methylationData.3.QC.mRNA %>% 
  group_by(Name,Tissue ) %>%
  summarise(depth = mean(depth), percent = mean(percent))->
  methylationData.3.QC.mRNA.Name.Tissue.summary

methylationData.3.QC.mRNA.Name.Tissue.summary %>%
  group_by(Name) %>%
  summarise(max = max(percent),min = min(percent)) %>%
  inner_join(methylationData.3.QC.mRNA.Name.Tissue.summary) %>%
  filter(percent == max) ->
  methylationData.3.QC.mRNA.Name.Tissue.summary.max 

  methylationData.3.QC.mRNA.Name.Tissue.summary.max %>%
  inner_join(kruskal_results.mRNA.sig) ->
    methylationData.3.QC.mRNA.Name.Tissue.summary.max.sig


methylationData.3.QC.mRNA.Name.Tissue.summary.max.sig %>%
  inner_join(annotation.mRNA.df) -> 
  methylationData.3.QC.mRNA.Name.Tissue.summary.max.sig


methylationData.3.QC.mRNA %>% filter(Name == "XM_043160398.1") %>%
  ggplot(aes(x = Tissue, y = percent)) + geom_violin() 

methylationData.3.QC.mRNA %>% filter(Name == "XM_043152787.1")  %>%
  ggplot(aes( x = start, y = percent, color = Tissue)) + geom_point(size = 0.3, aplpha = 0.5) +
  geom_smooth()



kruskal_results.mRNA.sig %>% slice_head(n = 5)  -> 
  top5.mRNA

methylationData.3.QC.mRNA   %>% 
  filter(Name %in%  top5.mRNA$Name) %>%
  mutate(Name = factor(Name, levels = top5.mRNA$Name)) %>%
  ggplot(aes(x = start, y = percent, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.4) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(0,100)) +
  labs(
    title = "Methylation across different TEs (top 5 significant)",
    x = "Position",
    y = "Percent methylated"
  ) + facet_wrap("Name", scales = "free_x")   +
  theme_light() -> examples.methylation
  

methylationData.3.QC.TE %>% 
  filter(TE_ID2 %in%  top5.te$TE_ID2) %>%
  ggplot(aes(x = start, y = depth, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.4) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(0,40))  +
  labs(
    title = "Depth across different TEs (top 5 significant)",
    x = "Position in TE",
    y = "Depth"
  ) + facet_wrap(c("TE_ID2","TEclass"),nrow = 1, scales = "free_x")   +
  theme_light()  -> examples.depth



```


### mRNA methylation levels

```{r }
methylationData.3.QC.mRNA %>%
  group_by(Name,Sample,Location,Replicate2,Tissue) %>%
  summarise( percent = mean(percent)) ->
  methylationData.3.QC.mRNA.Name.Tissue.summary


methylationData.3.QC.mRNA.Name.Tissue.summary %>%
  mutate(methylationLevel = "All") %>%
  mutate(methylationLevel = replace(methylationLevel, percent < 90, "High")) %>%
  mutate(methylationLevel = replace(methylationLevel, percent < 75 , "Medium"))%>%
  mutate(methylationLevel = replace(methylationLevel, percent < 30 , "Low"))%>%
  mutate(methylationLevel = replace(methylationLevel, percent < 5 , "Nothing"))  ->
methylationData.3.QC.mRNA.Name.Tissue.summary      
  
methylationData.3.QC.mRNA.Name.Tissue.summary %>%
  group_by(Sample, Tissue, methylationLevel) %>%
  summarise(n = n())


methylationData.3.QC.mRNA.Name.Tissue.summary %>%
  filter(methylationLevel == "High" | methylationLevel == "Medium" ) %>%
  group_by(Name) %>%
  summarise(
    percent = mean(percent),
    n = n()
  )  %>%
  filter(n == 26 )  %>%
  filter( percent > 60)  ->
  mRNA.high.methylation
  
         
  methylationData.3.QC.mRNA.Name.Tissue.summary %>%ggplot(aes(x = Sample, y = percent, fill = Tissue)) +
           geom_violin()


```


###  Both analysis

####  Non-Parametric Test (Kruskal-Wallis)
```{r }
  
methylationData.3.QC %>%
  filter(location == "Both")  ->
  methylationData.3.QC.Both




methylationData.3.QC.Both %>%
  group_by(Name) %>%
  summarise(
    p_value = tryCatch({
      kruskal.test(percent ~ Tissue, data = cur_data())$p.value
    }, error = function(e) NA)
  ) %>%
  mutate(significant = "No") %>%
  mutate(significant = replace(significant, p_value < 0.0000001, "Yes")) ->
  kruskal_results.Both



# Filter significant results
kruskal_results.Both %>%
  filter(!is.na(p_value) & p_value < 0.0000000001)%>%
  arrange(p_value) ->
  kruskal_results.Both.sig

methylationData.3.QC.Both %>% 
  group_by(Name,Tissue ) %>%
  summarise(depth = mean(depth), percent = mean(percent))->
  methylationData.3.QC.Both.Name.Tissue.summary

methylationData.3.QC.Both.Name.Tissue.summary %>%
  group_by(Name) %>%
  summarise(max = max(percent),min = min(percent)) %>%
  inner_join(methylationData.3.QC.Both.Name.Tissue.summary) %>%
  filter(percent == max) ->
  methylationData.3.QC.Both.Name.Tissue.summary.max 

  methylationData.3.QC.Both.Name.Tissue.summary.max %>%
  inner_join(kruskal_results.Both.sig) ->
    methylationData.3.QC.Both.Name.Tissue.summary.max.sig


methylationData.3.QC.Both.Name.Tissue.summary.max.sig %>%
  inner_join(annotation.mRNA.df) -> 
  methylationData.3.QC.Both.Name.Tissue.summary.max.sig


kruskal_results.Both.sig %>% slice_head(n = 5)  -> 
  top5.both

methylationData.3.QC.Both   %>% 
  filter(Name %in%  top5.both$Name) %>%
  mutate(Name = factor(Name, levels = top5.both$Name)) %>%
  ggplot(aes(x = start, y = percent, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.4) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(0,100)) +
  labs(
    title = "Methylation across different Both",
    x = "Position",
    y = "Percent methylated"
  ) + facet_wrap("Name", scales = "free_x")   +
  theme_light() -> examples.Both.methylation
  

methylationData.3.QC.TE %>% 
  filter(TE_ID2 %in%  top5.te$TE_ID2) %>%
  ggplot(aes(x = start, y = depth, color = Tissue)) +
  geom_point( size = 0.5, alpha = 0.4) + geom_smooth(se = FALSE) +
  scale_color_manual(values = colors) + ylim( c(0,40))  +
  labs(
    title = "Depth across different TEs (top 5 significant)",
    x = "Position in TE",
    y = "Depth"
  ) + facet_wrap(c("TE_ID2","TEclass"),nrow = 1, scales = "free_x")   +
  theme_light()  -> examples.depth





```


### Compare TE and mRNA
```{r }



# Place the TEs on the chromosomes.
methylationData.3.QC.TE.summary %>%
  inner_join(TEdf.long) -> methylationData.3.QC.TE.summary.location

# Keep only the long chromosomes
methylationData.3.QC.TE.summary.location %>%
  group_by(seqid) %>%
  summarise(n = n()) %>%
  filter(n > 50) %>%
  select(seqid)->
  chrs
  
chrs2 = chrs[1:11,]


# Rename so that mRNA and TE can be joined
methylationData.3.QC.TE.summary.location %>%
  select(Tissue, mean,  seqid, start) %>%
  mutate(type = "TE") ->
  ME.TE.location

methylationData.3.QC.mRNA.Name.Tissue.summary  %>%
  inner_join(annotation.mRNA.df) %>%
  filter(seqid %in% chrs2$seqid ) %>%
  select(Tissue, percent,  seqid, start) %>%
  dplyr::rename(mean = "percent") %>%
  mutate(type = "mRNA") ->
  ME.mRNA.location

bind_rows(ME.TE.location,ME.mRNA.location)->
  ME.TE.mRNA.location

ME.mRNA.location%>%
  filter(seqid %in% chrs2$seqid) %>%
  ggplot(aes(start,y =mean , color = Tissue)) + geom_point(size = 0.1) +
  facet_grid(Tissue~seqid) +
    theme_minimal()+
    scale_color_manual(values = colors) + ggtitle("mRNA methylation level") -> plot3

ME.TE.location %>%
  filter(seqid %in% chrs2$seqid) %>%
  ggplot(aes(start,y =mean , color = Tissue)) + geom_point(size = 0.1) +
  facet_grid(Tissue~seqid) +
    theme_minimal()+
    scale_color_manual(values = colors)+ ggtitle("TE methylation level")   -> plot1

ME.TE.mRNA.location %>% inner_join(TEdf.long) %>%
  filter(TEclass== "Gypsy") %>%
  filter(seqid %in% chrs2$seqid) %>%
  ggplot(aes(start,y =mean , color = type)) + geom_point(size = 0.1) +
  facet_grid(TEclass~seqid) +
    theme_minimal() ->plot2

ME.TE.mRNA.location %>% inner_join(TEdf.long) %>%
  filter(mean > 0.24) %>%
  filter(TEclass== "Gypsy") %>%
  filter(seqid %in% chrs2$seqid[1:3]) %>%
  ggplot(aes(start, color = type)) + geom_density(size = 0.5,bw = 50000) +
  facet_grid(TEclass~seqid) +
    theme_minimal() ->plot4

ME.TE.mRNA.location %>% inner_join(TEdf.long) %>%
  filter(mean < 0.24) %>%
  filter(TEclass== "Gypsy") %>%
  filter(seqid %in% chrs2$seqid) %>%
  ggplot(aes(start, color = type)) + geom_density(size = 0.5,bw = 50000) +
  facet_grid(TEclass~seqid) +
    theme_minimal() ->plot5

annotation.mRNA.df %>%
  filter(seqid %in% chrs2$seqid[1:3]) %>%
  ggplot(aes(start)) + geom_density(size = 0.3,bw = 50000) +
  facet_grid(.~seqid) +
    theme_minimal() -> plot6


plot3+plot5+plot4+plot1+ plot_layout(heights = c(2,1,1,2))
ggsave("../../results/meDNA/TE/TE.mRNA.location.pdf", width = 30, height = 10)

ME.TE.mRNA.location %>%
  filter(seqid %in% chrs2$seqid[1:3]) %>%
  ggplot(aes(start,y =mean , color = type)) + geom_point(size = 0.1) +
  facet_grid(Tissue~seqid) +
    theme_minimal()+ ggtitle(" Methylation level")  -> plot6


plot5+plot4+plot6+ plot_layout(heights = c(1,1,2))


ME.TE.location %>%
  filter(seqid %in% chrs2$seqid[11]) %>%
  ggplot(aes(start,y =mean , color = Tissue)) + geom_point(size = 0.1) +
  facet_grid(Tissue~seqid) +
    theme_minimal()+
    scale_color_manual(values = colors)+ ggtitle("TE methylation level")   -> plot1


 
  
  
  

```

```{r}

GOclass_counts <- kruskal_results.mRNA %>%
  inner_join(annotation.mRNA.GO.df) %>%
  group_by(GO, significant) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = significant, values_from = count, values_fill = 0) %>%
  rename(No = "No", Yes = "Yes") %>%
  mutate(ratio = No/Yes) %>%
  mutate(totalYes = sum(Yes)) %>%
  mutate(totalNo = sum(No)) %>%
  mutate(average =sum(Yes)/sum(No) )


GOclass_stats <- GOclass_counts %>%
  rowwise() %>%
  mutate(
    p_value = fisher.test(matrix(c(No, Yes, 8997, 29830), nrow = 2))$p.value
  ) %>%
  ungroup() %>% arrange(-ratio) %>%
  mutate(ratio = Yes/No) %>%
  filter(p_value < 0.00000001) %>%
  arrange(p_value)



```

```{r}

methylationData.3.QC.TE %>%
  ggplot(aes(x = Tissue, fill = methylationLevel ))+ geom_bar(position = "fill") +  # Replace with your desired geom
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) + facet_wrap("TEclass")


ggsave("../../results/meDNA/TE/MethylationDistribution.TEclass.pdf")





methylationData.3.QC.TE %>%ungroup() %>% 
  group_by(TE_ID2,Tissue) %>%
  filter(!is.na(percent)) %>%
  summarise(percent = mean(percent)) %>%
  pivot_wider(values_from = "percent",names_from = "Tissue") %>%
  ggplot(aes(x = spore, y = lamella)) + geom_point(size = 0.1)+ geom_smooth(method = "lm") +geom_density_2d_filled(alpha = 0.)

  

methylationData.3.QC.TE %>%
  ungroup() %>%
  dplyr::count(TE_ID2, Tissue) 

methylationData.3.QC.TE %>%
  summarise(has_na = any(is.na(percent)))

methylationData.3.QC.TE %>%
  ungroup() %>%
  group_by(TE_ID2, Tissue) %>%
  summarise(group_size = n())

```

```{r}



#methylationData.3 = read_csv("../../data/meDNA/methylationData.raw.annotated.csv")


methylationData.3.QC %>%
  filter(depth>10)%>% 
  ggplot(aes( x =log2(percent+1),
              color = Tissue,
              linetype = as.factor(Replicate))) +
  geom_density() +
  facet_grid(Location~location)





methylationData.3.wide %>% column_to_rownames(var = "posID") -> 
  methylationData.3.wide

```

#### Coding gene analysis
```{r }

methylationData.3.QC %>%
  filter(location == "GeneBody")->
  MD.mRNA.QC


samples = MD.mRNA.QC %>% ungroup() %>% distinct(Sample)

mRNA.data = list()

for(i in 1:nrow(samples)){
  MD.mRNA.QC %>%
    filter(Sample == samples$Sample[[i]]) ->
    MD.mRNA.QC.sample
  MD.mRNA.QC.sample%>%
  group_by(Name) %>%
  summarise(depth = mean(depth), percent = mean(percent)) %>%
    mutate(Sample = samples$Sample[[i]]) ->
  MD.mRNA.QC.sample.summary
  mRNA.data[[samples$Sample[[i]]]] = MD.mRNA.QC.sample.summary
}

MD.mRNA.summary = bind_rows(mRNA.data)


MD.mRNA.summary %>%
  group_by(Sample) %>%
  summarise(n = n())


MD.mRNA.summary %>%
  group_by(Name) %>%
  summarise(n = n()) %>%
  filter(n == 26) %>% 
  inner_join(MD.mRNA.summary) ->
  MD.mRNA.summary.QC


MD.mRNA.summary.QC %>% inner_join(samples.ring.methyl.df) -> 
  MD.mRNA.summary.QC

MD.mRNA.summary.QC %>% group_by(Name,Tissue ) %>%
  summarise(mean = mean(percent) , sd =sd(percent) ) ->
  MD.mRNA.summary.QC.summary

MD.mRNA.summary.QC.summary %>% group_by(Name) %>%
  summarise(mean = mean(mean) , sd =sd(mean) ) %>%
  arrange(-sd)


MD.mRNA.summary.QC.summary %>% select(Name,Tissue,mean) %>%
  pivot_wider(names_from = "Tissue",values_from = "mean") ->
  MD.mRNA.summary.QC.wide

ggplot(MD.mRNA.summary.QC.wide, aes(x = lamella, y = stipe)) + 
  geom_point()


```



## Spore methylation pattern (check litterature)

## Check coverage of different TE for getting information about abundance of TEs
Probing TE data for the different classes if they have increased or decreased in the different tissues. 
```{}



```

## Check TE expression 
Get epxression levels for the 
```{}


```


 


